### 2024-01-19 outputs/paper/paper.qmd
Added structure to repo with initial files

---
title: "(Working Title) Motor Vehicle Collision and Income of Wards"
subtitle: "My subtitle if needed"
author: "Emily Su"
thanks: "Data and code are available at: https://github.com/moonsdust/toronto-collisions"
date: today
date-format: "D MMMM YYYY"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("outputs/models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-01-19 scripts/00-simulate_data.R

@@ -1,0 +1,19 @@
+#### Preamble ####
+# Purpose: Simulates... [...UPDATE THIS...]
+# Author: Emily Su
+# Date: 25 January 2024
+# Contact: em.su@utoronto.ca
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+# [...UPDATE THIS...]
+
+#### Simulate data ####
+# [...ADD CODE HERE...]
+
+
+

### 2024-01-19 scripts/01-download_data.R

@@ -1,0 +1,26 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from [...UPDATE THIS...]
+# Author: Emily Su
+# Date: 25 January 2024
+# Contact: em.su@utoronto.ca
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(opendatatoronto)
+library(tidyverse)
+# [...UPDATE THIS...]
+
+#### Download data ####
+# [...ADD CODE HERE TO DOWNLOAD...]
+
+
+
+#### Save data ####
+# [...UPDATE THIS...]
+# change the_raw_data to whatever name you assigned when you downloaded it.
+write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+
+         

### 2024-01-19 scripts/02-data_cleaning.R

@@ -1,0 +1,44 @@
+#### Preamble ####
+# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
+# Author: Emily Su
+# Date: 25 January 2024
+# Contact: em.su@utoronto.ca
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+#### Workspace setup ####
+library(tidyverse)
+
+#### Clean data ####
+raw_data <- read_csv("inputs/data/plane_data.csv")
+
+cleaned_data <-
+  raw_data |>
+  janitor::clean_names() |>
+  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
+  filter(wing_width_mm != "caw") |>
+  mutate(
+    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
+                                   "1.35",
+                                   flying_time_sec_first_timer)
+  ) |>
+  mutate(wing_width_mm = if_else(wing_width_mm == "490",
+                                 "49",
+                                 wing_width_mm)) |>
+  mutate(wing_width_mm = if_else(wing_width_mm == "6",
+                                 "60",
+                                 wing_width_mm)) |>
+  mutate(
+    wing_width_mm = as.numeric(wing_width_mm),
+    wing_length_mm = as.numeric(wing_length_mm),
+    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
+  ) |>
+  rename(flying_time = flying_time_sec_first_timer,
+         width = wing_width_mm,
+         length = wing_length_mm
+         ) |> 
+  tidyr::drop_na()
+
+#### Save data ####
+write_csv(cleaned_data, "outputs/data/analysis_data.csv")

### 2024-01-19 scripts/03-test_data.R

@@ -1,0 +1,15 @@
+#### Preamble ####
+# Purpose: Tests... [...UPDATE THIS...]
+# Author: Emily Su
+# Date: 25 January 2024
+# Contact: em.su@utoronto.ca
+# License: MIT
+# Pre-requisites: [...UPDATE THIS...]
+# Any other information needed? [...UPDATE THIS...]
+
+
+#### Workspace setup ####
+library(tidyverse)
+# [...UPDATE THIS...]
+
+#### Test data ####

### 2024-01-19 outputs/paper/paper.qmd
updated title

@@
 ---
-title: "(Working Title) Motor Vehicle Collision and Income of Wards"
+title: "(Working Title) Motor Vehicle Collision Across Toronto"
 subtitle: "My subtitle if needed"

### 2024-01-19 outputs/paper/paper.qmd
updated title of paper

@@
 ---
-title: "(Working Title) Motor Vehicle Collision Across Toronto"
+title: "(Working Title) Motor Vehicle Collisions Across Toronto"
 subtitle: "My subtitle if needed"

### 2024-01-20 scripts/00-simulate_data.R
Finished script to simulate datasets on collisions in Toronto

@@ -1,19 +1,103 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
+# Purpose: To simulate datasets regarding motor vehicle collisions 
+# across Toronto 
 # Author: Emily Su
 # Date: 25 January 2024
 # Contact: em.su@utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: -
 
+#### Workspace setup #### 
 
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
+## Installing packages (only needs to be done once per computer)
+# Uncomment the below if this is the first time running the script 
+# install.packages("tidyverse")
+# install.packages("knitr")
+library(tidyverse) # Contains data-related packages
+library(knitr) # To make tables
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(646)
+
+### Global Variables ###
+number_of_years <- 7 
+num_of_wards <- 25 # Number of wards in Toronto
+toronto_pop <- 2930000 # Population of Toronto
+
+## Dataset 1: Simulate dataset regarding the number of collisions, 
+# collision types, and what year the collision occured ##
+# Expected Columns: year | collision_type | num_of_collisions 
+simulated_data_one <-
+  tibble(
+    # We will repeat the type of collision name for <number_of_years> times 
+    collision_type =
+      c(rep("Personal Injury", number_of_years), 
+        rep("Fail to Remain", number_of_years), 
+        rep("Property Damage", number_of_years), 
+        rep("Fatal", number_of_years)),
+    # We will create a vector for the years 2017 to 2023
+    # and repeat this 4 times for each type of collision 
+    year = rep(c(1:number_of_years + 2016), 4),
+    # We will create <number_of_years> * 4 random values 
+    # from a uniform distribution to get the number of collisions 
+    # each year for each collision type 
+    num_of_collisions =
+      round(runif(n = number_of_years * 4, min = 0, max = 1000))
+  )
+# Creates table for the simulated data
+simulated_data_one
+
+# Dataset 1: Data Validation / Tests
+# 1. "year" does not contain years before 2017 and after 2023
+simulated_data_one$year |> min() == 2017
+simulated_data_one$year |> max() == 2023
+
[…]

### 2024-01-20 scripts/00-simulate_data.R
Updated prereqs in 00-simulate_data.R

@@ -5,7 +5,8 @@
 # Date: 25 January 2024
 # Contact: em.su@utoronto.ca
 # License: MIT
-# Pre-requisites: -
+# Pre-requisites: tidyverse and knitr have been installed, else 
+# uncomment the code to install the two packages.
 
 #### Workspace setup #### 
 

### 2024-01-20 scripts/00-install_packages.R
Created script to install packages for workspace

@@ -1,0 +1,13 @@
+#### Preamble ####
+# Purpose: Installs packages needed to run scripts and Quarto document
+# Author: Emily Su
+# Date: 25 January 2024
+# Contact: em.su@utoronto.ca
+# License: MIT
+# Pre-requisites: -
+
+#### Workspace setup ####
+## Installing packages (only needs to be done once per computer)
+install.packages("tidyverse") # Contains data-related packages
+install.packages("knitr") # To make tables
+install.packages("opendatatoronto")

### 2024-01-20 scripts/00-simulate_data.R

@@ -5,17 +5,11 @@
 # Date: 25 January 2024
 # Contact: em.su@utoronto.ca
 # License: MIT
-# Pre-requisites: tidyverse and knitr have been installed, else 
-# uncomment the code to install the two packages.
+# Pre-requisites: tidyverse has been installed by running 00-install_packages.R
 
 #### Workspace setup #### 
-
-## Installing packages (only needs to be done once per computer)
-# Uncomment the below if this is the first time running the script 
-# install.packages("tidyverse")
-# install.packages("knitr")
+# Loading packages 
 library(tidyverse) # Contains data-related packages
-library(knitr) # To make tables
 
 #### Simulate data ####
 set.seed(646)
@@ -24,6 +18,8 @@ set.seed(646)
 number_of_years <- 7 
 num_of_wards <- 25 # Number of wards in Toronto
 toronto_pop <- 2930000 # Population of Toronto
+
+
 
 ## Dataset 1: Simulate dataset regarding the number of collisions, 
 # collision types, and what year the collision occured ##

### 2024-01-20 scripts/00-install_packages.R
Updated scripts with the correct preamble info

@@ -2,7 +2,7 @@
 # Purpose: Installs packages needed to run scripts and Quarto document
 # Author: Emily Su
 # Date: 25 January 2024
-# Contact: em.su@utoronto.ca
+# Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: -
 

### 2024-01-20 scripts/00-simulate_data.R

@@ -3,7 +3,7 @@
 # across Toronto 
 # Author: Emily Su
 # Date: 25 January 2024
-# Contact: em.su@utoronto.ca
+# Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: tidyverse has been installed by running 00-install_packages.R
 

### 2024-01-20 scripts/01-download_data.R

@@ -1,20 +1,24 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from Open Data Toronto 
+# on traffic collisions and information about the Toronto wards.  
 # Author: Emily Su
 # Date: 25 January 2024
-# Contact: em.su@utoronto.ca
+# Contact: em.su@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: opendatatoronto and tidyverse has been installed 
+# by running 00-install_packages.R
 
+# Datasets that will be downloaded and saved:
+# 1. Police Annual Statistical Report - Traffic Collisions: https://open.toronto.ca/dataset/police-annual-statistical-report-traffic-collisions/
+# 2. City Wards - Traffic Collisions: https://open.toronto.ca/dataset/city-wards/
+# 3. Ward Profiles (25-Ward Model): https://open.toronto.ca/dataset/ward-profiles-25-ward-model/
 
 #### Workspace setup ####
+# Loading packages
 library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
 
 
 

### 2024-01-20 scripts/03-test_data.R

@@ -2,7 +2,7 @@
 # Purpose: Tests... [...UPDATE THIS...]
 # Author: Emily Su
 # Date: 25 January 2024
-# Contact: em.su@utoronto.ca
+# Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: [...UPDATE THIS...]
 # Any other information needed? [...UPDATE THIS...]

### 2024-01-20 scripts/01-download_data.R
Finished script to download datasets and uploaded raw data

@@ -7,24 +7,45 @@
 # License: MIT
 # Pre-requisites: opendatatoronto and tidyverse has been installed 
 # by running 00-install_packages.R
+# IMPORTANT NOTE: The "Police Annual Statistical Report - Traffic Collisions" dataset, 
+# is a large file so it might take a while to finish downloading and saving the file. 
 
 # Datasets that will be downloaded and saved:
-# 1. Police Annual Statistical Report - Traffic Collisions: https://open.toronto.ca/dataset/police-annual-statistical-report-traffic-collisions/
+# 1. Ward Profiles (25-Ward Model): https://open.toronto.ca/dataset/ward-profiles-25-ward-model/
 # 2. City Wards - Traffic Collisions: https://open.toronto.ca/dataset/city-wards/
-# 3. Ward Profiles (25-Ward Model): https://open.toronto.ca/dataset/ward-profiles-25-ward-model/
+# 3. Police Annual Statistical Report - Traffic Collisions: https://open.toronto.ca/dataset/police-annual-statistical-report-traffic-collisions/
 
 #### Workspace setup ####
 # Loading packages
 library(opendatatoronto)
 library(tidyverse)
 
-#### Download data ####
 
+#### Download and Save data ####
+# Code snippet to filter for and download datasets was adapted from: 
+# https://open.toronto.ca/dataset/police-annual-statistical-report-traffic-collisions/
 
+# Function to download files
+download_csv_geojson <- function(files) {
+  filter(files, row_number()==1) |> get_resource()
+}
 
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+# Download "Ward Profiles (25-Ward Model)" dataset
+ward_profiles_data <- download_csv_geojson(list_package_resources("6678e1a6-d25f-4dff-b2b7-aa8f042bc2eb"))
+# Save "Ward Profiles (25-Ward Model)" dataset
+write_csv(as.data.frame(ward_profiles_data[1]), "inputs/data/unedited_ward_profiles_data.csv")
 
-         
+# Download "City Wards" dataset
+# Filter for CSV and geojson files in the package resources list
+files_one <- filter(list_package_resources("5e7a8234-f805-43ac-820f-03d7c360b588"), tolower(format) %in% c('csv', 'geojson'))
[…]

### 2024-01-20 scripts/02-data_cleaning.R
Fixed spelling in simulate data script

@@ -1,16 +1,19 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
+# Purpose: Cleans unedited collisions data and unedited ward profiles data
+# to create clean, usable CSV files with relevant columns.
 # Author: Emily Su
 # Date: 25 January 2024
-# Contact: em.su@utoronto.ca
+# Contact: em.su@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: tidyverse has been installed by running 00-install_packages.R
+# and datasets have been downloaded by running 01-download_data.R. 
 
 #### Workspace setup ####
 library(tidyverse)
 
 #### Clean data ####
+
+# Dataset 1 (Expected Columns)
 raw_data <- read_csv("inputs/data/plane_data.csv")
 
 cleaned_data <-

### 2024-01-20 scripts/00-simulate_data.R
Fixed spelling in simulate data script

@@ -22,7 +22,7 @@ toronto_pop <- 2930000 # Population of Toronto
 
 
 ## Dataset 1: Simulate dataset regarding the number of collisions, 
-# collision types, and what year the collision occured ##
+# collision types, and what year the collision occurred ##
 # Expected Columns: year | collision_type | num_of_collisions 
 simulated_data_one <-
   tibble(

### 2024-01-20 scripts/00-install_packages.R
added call to install janitor package

@@ -11,3 +11,4 @@
 install.packages("tidyverse") # Contains data-related packages
 install.packages("knitr") # To make tables
 install.packages("opendatatoronto")
+install.packages("janitor")

### 2024-01-21 scripts/00-install_packages.R
added dplyr to install packages script

@@ -12,3 +12,4 @@ install.packages("tidyverse") # Contains data-related packages
 install.packages("knitr") # To make tables
 install.packages("opendatatoronto")
 install.packages("janitor")
+install.packages("dplyr")

### 2024-01-21 scripts/00-install_packages.R
added ggplot2 to install packages script

@@ -11,5 +11,6 @@
 install.packages("tidyverse") # Contains data-related packages
 install.packages("knitr") # To make tables
 install.packages("opendatatoronto")
-install.packages("janitor")
+install.packages("janitor") # To clean datasets
 install.packages("dplyr")
+install.packages(ggplot2) # To make graphs

### 2024-01-21 scripts/00-install_packages.R
added in double quotes in call to install package

@@ -13,4 +13,4 @@ install.packages("knitr") # To make tables
 install.packages("opendatatoronto")
 install.packages("janitor") # To clean datasets
 install.packages("dplyr")
-install.packages(ggplot2) # To make graphs
+install.packages("ggplot2") # To make graphs

### 2024-01-21 scripts/00-simulate_data.R
Removed Fail to Remain Collision Type from Simulation Script

@@ -28,18 +28,17 @@ simulated_data_one <-
   tibble(
     # We will repeat the type of collision name for <number_of_years> times 
     collision_type =
-      c(rep("Personal Injury", number_of_years), 
-        rep("Fail to Remain", number_of_years), 
+      c(rep("Personal Injury", number_of_years),
         rep("Property Damage", number_of_years), 
         rep("Fatal", number_of_years)),
     # We will create a vector for the years 2017 to 2023
-    # and repeat this 4 times for each type of collision 
-    year = rep(c(1:number_of_years + 2016), 4),
-    # We will create <number_of_years> * 4 random values 
+    # and repeat this 3 times for each type of collision 
+    year = rep(c(1:number_of_years + 2016), 3),
+    # We will create <number_of_years> * 3 random values 
     # from a uniform distribution to get the number of collisions 
     # each year for each collision type 
     num_of_collisions =
-      round(runif(n = number_of_years * 4, min = 0, max = 1000))
+      round(runif(n = number_of_years * 3, min = 0, max = 1000))
   )
 # Creates table for the simulated data
 simulated_data_one
@@ -49,14 +48,14 @@ simulated_data_one
 simulated_data_one$year |> min() == 2017
 simulated_data_one$year |> max() == 2023
 
-# 2. We have 4 types of collisions: "Personal Injury", "Fail to Remain", 
+# 2. We have 3 types of collisions: "Personal Injury",
 # "Property Damage", "Fatal"
 simulated_data_one$collision_type |>
   unique() |>
-  length() == 4
+  length() == 3
 
 simulated_data_one$collision_type |>
-  unique() == c("Personal Injury", "Fail to Remain", "Property Damage", "Fatal")
+  unique() == c("Personal Injury", "Property Damage", "Fatal")
 
 # 3. num_of_collisions is greater than or equal to 0.
 simulated_data_one$num_of_collisions |> min() >= 0

### 2024-01-21 scripts/00-simulate_data.R
added back in fail to remain type and updated sketch 4

@@ -28,17 +28,18 @@ simulated_data_one <-
   tibble(
     # We will repeat the type of collision name for <number_of_years> times 
     collision_type =
-      c(rep("Personal Injury", number_of_years),
+      c(rep("Personal Injury", number_of_years), 
+        rep("Fail to Remain", number_of_years), 
         rep("Property Damage", number_of_years), 
         rep("Fatal", number_of_years)),
     # We will create a vector for the years 2017 to 2023
-    # and repeat this 3 times for each type of collision 
-    year = rep(c(1:number_of_years + 2016), 3),
-    # We will create <number_of_years> * 3 random values 
+    # and repeat this 4 times for each type of collision 
+    year = rep(c(1:number_of_years + 2016), 4),
+    # We will create <number_of_years> * 4 random values 
     # from a uniform distribution to get the number of collisions 
     # each year for each collision type 
     num_of_collisions =
-      round(runif(n = number_of_years * 3, min = 0, max = 1000))
+      round(runif(n = number_of_years * 4, min = 0, max = 1000))
   )
 # Creates table for the simulated data
 simulated_data_one
@@ -48,14 +49,14 @@ simulated_data_one
 simulated_data_one$year |> min() == 2017
 simulated_data_one$year |> max() == 2023
 
-# 2. We have 3 types of collisions: "Personal Injury",
+# 2. We have 4 types of collisions: "Personal Injury", "Fail to Remain", 
 # "Property Damage", "Fatal"
 simulated_data_one$collision_type |>
   unique() |>
-  length() == 3
+  length() == 4
 
 simulated_data_one$collision_type |>
-  unique() == c("Personal Injury", "Property Damage", "Fatal")
+  unique() == c("Personal Injury", "Fail to Remain", "Property Damage", "Fatal")
 
 # 3. num_of_collisions is greater than or equal to 0.
 simulated_data_one$num_of_collisions |> min() >= 0

### 2024-01-21 scripts/00-install_packages.R
Updated sketch 4 and date on scripts

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Installs packages needed to run scripts and Quarto document
 # Author: Emily Su
-# Date: 25 January 2024
+# Date: 23 January 2024
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: -

### 2024-01-21 scripts/00-simulate_data.R

@@ -2,7 +2,7 @@
 # Purpose: To simulate datasets regarding motor vehicle collisions 
 # across Toronto 
 # Author: Emily Su
-# Date: 25 January 2024
+# Date: 23 January 2024
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: tidyverse has been installed by running 00-install_packages.R

### 2024-01-21 scripts/01-download_data.R

@@ -2,7 +2,7 @@
 # Purpose: Downloads and saves the data from Open Data Toronto 
 # on traffic collisions and information about the Toronto wards.  
 # Author: Emily Su
-# Date: 25 January 2024
+# Date: 23 January 2024
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: opendatatoronto and tidyverse has been installed 

### 2024-01-21 scripts/03-test_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests... [...UPDATE THIS...]
 # Author: Emily Su
-# Date: 25 January 2024
+# Date: 23 January 2024
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: [...UPDATE THIS...]

### 2024-01-21 scripts/02-data_cleaning.R
Finished data cleaning script, uploaded data, and updated sketches

@@ -2,7 +2,7 @@
 # Purpose: Cleans unedited collisions data and unedited ward profiles data
 # to create clean, usable CSV files with relevant columns.
 # Author: Emily Su
-# Date: 25 January 2024
+# Date: 23 January 2024
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: tidyverse has been installed by running 00-install_packages.R
@@ -10,38 +10,190 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
+library(dplyr)
 
 #### Clean data ####
 
-# Dataset 1 (Expected Columns)
-raw_data <- read_csv("inputs/data/plane_data.csv")
+## Read in datasets ##
+raw_collisions_data <- read_csv("inputs/data/unedited_collisions_data.csv", show_col_types = FALSE)
+raw_ward_profiles_data <- read_csv("inputs/data/unedited_ward_profiles_data.csv", show_col_types = FALSE)
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
+## For collisions dataset ##
+# Clean the column names for collision data
+cleaned_collisions_data <- clean_names(raw_collisions_data)
+
+# Select our desired columns from this dataset that we will want for 
+# the new, cleaned dataset 
+cleaned_collisions_data <-
+  cleaned_collisions_data |>
+  filter(occ_year >= 2017 & occ_year <= 2023) |>
+  select(
+    occ_year,
+    fatalities,
+    injury_collisions,
+    pd_collisions,
+    ftr_collisions, 
+    long_wgs84,
+    lat_wgs84
+  )
+
+# Renaming occ_year column to year
+cleaned_collisions_data <-
+  cleaned_collisions_data |>
+  rename( # Left side is the new name and the right side is the old name
+    year = occ_year
+  )
+
+# Get rid of all rows with NA 
+cleaned_collisions_data <- na.omit(cleaned_collisions_data)
+
+# Replace NO with 0 and YES with 1 and convert to numeric
+cleaned_collisions_data$injury_collisions <- str_replace(
[…]

### 2024-01-21 scripts/02-data_cleaning.R
Fixed ordering of columns in cleaned_ward_data.csv

@@ -157,7 +157,8 @@ ward_pop_size$pop_num <- as.numeric(ward_pop_size$pop_num)
 # Join ward_pop_size, and ward_avg_income_row into a single tibble 
 cleaned_ward_data <- 
   ward_pop_size |>
-  right_join(ward_avg_income, by = "ward_num")
+  right_join(ward_avg_income, by = "ward_num") |>
+  select(ward_num, pop_num, avg_income) # Rearrange order
 
 
 # Dataset 3 (Contains data of collision type and their longitude and latitude)

### 2024-01-21 scripts/00-install_packages.R
added rgdal

@@ -14,3 +14,4 @@ install.packages("opendatatoronto")
 install.packages("janitor") # To clean datasets
 install.packages("dplyr")
 install.packages("ggplot2") # To make graphs
+install.packages("rgdal") 

### 2024-01-21 scripts/00-install_packages.R
removed rgdal

@@ -14,4 +14,3 @@ install.packages("opendatatoronto")
 install.packages("janitor") # To clean datasets
 install.packages("dplyr")
 install.packages("ggplot2") # To make graphs
-install.packages("rgdal") 

### 2024-01-21 scripts/00-install_packages.R
Added sf package to install packages script

@@ -14,3 +14,5 @@ install.packages("opendatatoronto")
 install.packages("janitor") # To clean datasets
 install.packages("dplyr")
 install.packages("ggplot2") # To make graphs
+install.packages("sf") # For making maps
+

### 2024-01-21 scripts/02-data_cleaning.R
Updated data cleaning script to omit 0s in coordinates

@@ -189,6 +189,9 @@ dataset_3 <-
   rename ( # Left side is the new name and the right side is the old name
     long = long_wgs84,
     lat = lat_wgs84
+  ) |>
+  filter(
+    (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
   )
 
 #### Save data ####

### 2024-01-21 scripts/00-install_packages.R
Added sf library

@@ -15,4 +15,3 @@ install.packages("janitor") # To clean datasets
 install.packages("dplyr")
 install.packages("ggplot2") # To make graphs
 install.packages("sf") # For making maps
-

### 2024-01-21 scripts/02-data_cleaning.R
Added neighbourhood column to dataset 3 for map data

@@ -35,7 +35,8 @@ cleaned_collisions_data <-
     pd_collisions,
     ftr_collisions, 
     long_wgs84,
-    lat_wgs84
+    lat_wgs84, 
+    neighbourhood_158
   )
 
 # Renaming occ_year column to year
@@ -75,7 +76,7 @@ cleaned_collisions_data$ftr_collisions <- as.numeric(cleaned_collisions_data$ftr
 # Create a new dataset
 dataset_1 <-
   cleaned_collisions_data |>
-  select(-c(long_wgs84,lat_wgs84)) |>
+  select(-c(long_wgs84,lat_wgs84, neighbourhood_158)) |>
   group_by(year, fatalities, injury_collisions, pd_collisions, ftr_collisions) |>
   mutate(num = n()) |> # count number of rows based on premise and year
   rename(num_of_collisions = num) |>
@@ -162,7 +163,7 @@ cleaned_ward_data <-
 
 
 # Dataset 3 (Contains data of collision type and their longitude and latitude)
-# Expected: collision_type | long | lat
+# Expected: collision_type | long | lat | neighbourhood
 dataset_3 <- 
   cleaned_collisions_data |>
   unique() |> # This will filter out repeated rows  
@@ -185,10 +186,11 @@ dataset_3 <-
       )
   ) |>
   select(-c(year, fatalities, injury_collisions, pd_collisions, ftr_collisions)) |> # Remove columns
-  select(collision_type, long_wgs84, lat_wgs84) |> # Rearrange columns in desired order  
+  select(collision_type, long_wgs84, lat_wgs84, neighbourhood_158) |> # Rearrange columns in desired order  
   rename ( # Left side is the new name and the right side is the old name
     long = long_wgs84,
-    lat = lat_wgs84
+    lat = lat_wgs84,
+    neighbourhood = neighbourhood_158
   ) |>
   filter(
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 

### 2024-01-22 scripts/02-data_cleaning.R
Added num_of_collisions column to dataset 3 for map data

@@ -162,8 +162,10 @@ cleaned_ward_data <-
   select(ward_num, pop_num, avg_income) # Rearrange order
 
 
-# Dataset 3 (Contains data of collision type and their longitude and latitude)
-# Expected: collision_type | long | lat | neighbourhood
+# Dataset 3 (Contains data of collision type, their longitude and latitude, 
+# neighbourhood of collision, and number of collisions of each neighbourhood
+# from 2017 to 2023)
+# Expected: collision_type | long | lat | neighbourhood | num_of)
 dataset_3 <- 
   cleaned_collisions_data |>
   unique() |> # This will filter out repeated rows  
@@ -194,7 +196,11 @@ dataset_3 <-
   ) |>
   filter(
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
-  )
+  ) |>
+  group_by(neighbourhood) |>
+  mutate(num = n()) |> # count number of rows based on premise and year
+  rename(num_of_collisions = num) |>
+  unique() # This will filter out repeated rows
 
 #### Save data ####
 # Save dataset 1

### 2024-01-22 scripts/02-data_cleaning.R
Grouped by collision_type for dataset 3

@@ -197,7 +197,7 @@ dataset_3 <-
   filter(
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
   ) |>
-  group_by(neighbourhood) |>
+  group_by(neighbourhood, collision_type) |>
   mutate(num = n()) |> # count number of rows based on premise and year
   rename(num_of_collisions = num) |>
   unique() # This will filter out repeated rows

### 2024-01-22 scripts/02-data_cleaning.R
Reverting previous commit

@@ -197,7 +197,7 @@ dataset_3 <-
   filter(
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
   ) |>
-  group_by(neighbourhood, collision_type) |>
+  group_by(neighbourhood) |>
   mutate(num = n()) |> # count number of rows based on premise and year
   rename(num_of_collisions = num) |>
   unique() # This will filter out repeated rows

### 2024-01-22 scripts/02-data_cleaning.R
Updated comment in 02-data_cleaning.R regarding mutate

@@ -198,7 +198,7 @@ dataset_3 <-
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
   ) |>
   group_by(neighbourhood) |>
-  mutate(num = n()) |> # count number of rows based on premise and year
+  mutate(num = n()) |> # count number of rows based on neighbourhood
   rename(num_of_collisions = num) |>
   unique() # This will filter out repeated rows
 

### 2024-01-22 scripts/02-data_cleaning.R
Included yearly collision and total collision columns

@@ -165,7 +165,7 @@ cleaned_ward_data <-
 # Dataset 3 (Contains data of collision type, their longitude and latitude, 
 # neighbourhood of collision, and number of collisions of each neighbourhood
 # from 2017 to 2023)
-# Expected: collision_type | long | lat | neighbourhood | num_of)
+# Expected: year | collision_type | long | lat | neighbourhood | num_of_collision | yearly_collision_num)
 dataset_3 <- 
   cleaned_collisions_data |>
   unique() |> # This will filter out repeated rows  
@@ -187,8 +187,8 @@ dataset_3 <-
         TRUE ~ "None"
       )
   ) |>
-  select(-c(year, fatalities, injury_collisions, pd_collisions, ftr_collisions)) |> # Remove columns
-  select(collision_type, long_wgs84, lat_wgs84, neighbourhood_158) |> # Rearrange columns in desired order  
+  select(-c(fatalities, injury_collisions, pd_collisions, ftr_collisions)) |> # Remove columns
+  select(year, collision_type, long_wgs84, lat_wgs84, neighbourhood_158) |> # Rearrange columns in desired order  
   rename ( # Left side is the new name and the right side is the old name
     long = long_wgs84,
     lat = lat_wgs84,
@@ -197,10 +197,18 @@ dataset_3 <-
   filter(
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
   ) |>
-  group_by(neighbourhood) |>
-  mutate(num = n()) |> # count number of rows based on neighbourhood
+  group_by(neighbourhood, collision_type) |>
+  mutate(num = n()) |> # count number of rows based on neighbourhood and collision type
   rename(num_of_collisions = num) |>
-  unique() # This will filter out repeated rows
+  unique() |> # This will filter out repeated rows
+  group_by(neighbourhood, year) |>
+  mutate(num = n()) |> # count number of rows based on neighbourhood and year
+  rename(yearly_collision_num = num) |>
+  unique() |>
+  group_by(neighbourhood) |>
+  mutate(num = n()) |>
+  rename(total_collisions_2017_2023 = num) |>
+  unique()
 
 #### Save data ####
 # Save dataset 1

### 2024-01-22 scripts/02-data_cleaning.R
Updated cleaned_ward_data.csv

@@ -135,7 +135,7 @@ ward_pop_size <-
 # Get row that contains the average household income
 ward_avg_income_row <- 
   cleaned_ward_data |> 
-  filter(x2021_one_variable_city_of_toronto_profiles == "Average total income of households in 2020 ($)")
+  filter(x2021_one_variable_city_of_toronto_profiles == "Total - Household total income groups in 2020 for private households - 25% sample data")
 # Convert row to a column 
 ward_avg_income <- 
   ward_avg_income_row |>

### 2024-01-22 scripts/02-data_cleaning.R
Updated cleaned_ward_data.csv to include ward names

@@ -129,7 +129,32 @@ ward_pop_size <-
     pop_num = "c(t(ward_pop_size_row))"
   ) |>
   mutate(
-    ward_num = c(1:25)
+    ward_num = c(1:25), 
+    ward_name = c("Etobicoke North", 
+                   "Etobicoke Centre", 
+                   "Etobicoke-Lakeshore", 
+                   "Parkdale-High Park", 
+                   "York South-Weston", 
+                   "York Centre", 
+                   "Humber River-Black Creek",
+                   "Eglinton-Lawrence", 
+                   "Davenport", 
+                   "Spadina-Fort York", 
+                   "University-Rosedale", 
+                   "Toronto-St. Paul’s", 
+                   "Toronto Centre", 
+                   "Toronto-Danforth", 
+                   "Don Valley West", 
+                   "Don Valley East",
+                   "Don Valley North", 
+                   "Willowdale", 
+                   "Beaches-East York", 
+                   "Scarborough Southwest", 
+                   "Scarborough Centre",
+                   "Scarborough-Agincourt",
+                   "Scarborough North", 
+                   "Scarborough-Guildwood",
+                   "Scarborough-Rouge Park")
   )
 
 # Get row that contains the average household income
@@ -148,7 +173,32 @@ ward_avg_income <-
     avg_income = "c(t(ward_avg_income_row))"
   ) |> 
   mutate(
-    ward_num = c(1:25)
+    ward_num = c(1:25), 
+    ward_name_avg = c("Etobicoke North", 
+                   "Etobicoke Centre", 
+                   "Etobicoke-Lakeshore", 
+                   "Parkdale-High Park", 
+                   "York South-Weston", 
+                   "York Centre", 
+                   "Humber River-Black Creek",
+                   "Eglinton-Lawrence", 
+                   "Davenport", 
+                   "Spadina-Fort York", 
+                   "University-Rosedale", 
+                   "Toronto-St. Paul’s", 
+                   "Toronto Centre", 
+                   "Toronto-Danforth", 
+                   "Don Valley West", 
+                   "Don Valley East",
+                   "Don Valley North", 
+                   "Willowdale", 
+                   "Beaches-East York", 
+                   "Scarborough Southwest", 
+                   "Scarborough Centre",
+                   "Scarborough-Agincourt",
+                   "Scarborough North", 
[…]

### 2024-01-22 scripts/00-simulate_data.R
Finished script to test cleaned datasets

@@ -56,7 +56,7 @@ simulated_data_one$collision_type |>
   length() == 4
 
 simulated_data_one$collision_type |>
-  unique() == c("Personal Injury", "Fail to Remain", "Property Damage", "Fatal")
+  unique() %in% c("Personal Injury", "Fail to Remain", "Property Damage", "Fatal")
 
 # 3. num_of_collisions is greater than or equal to 0.
 simulated_data_one$num_of_collisions |> min() >= 0

### 2024-01-22 scripts/02-data_cleaning.R

@@ -109,7 +109,7 @@ dataset_1 <-
 
 
 # Dataset 2 (cleaned_ward_data)
-# Expected columns: ward_num | pop_num | avg_income
+# Expected columns: ward_num | ward_name | pop_num | avg_income
 # Clean current column names
 cleaned_ward_data <- clean_names(raw_ward_profiles_data)
 
@@ -215,7 +215,8 @@ cleaned_ward_data <-
 # Dataset 3 (Contains data of collision type, their longitude and latitude, 
 # neighbourhood of collision, and number of collisions of each neighbourhood
 # from 2017 to 2023)
-# Expected: year | collision_type | long | lat | neighbourhood | num_of_collision | yearly_collision_num)
+# Expected: year | collision_type | long | lat | neighbourhood | 
+# num_of_collisions | yearly_collision_num | total_collisions_2017_2023
 dataset_3 <- 
   cleaned_collisions_data |>
   unique() |> # This will filter out repeated rows  
@@ -249,15 +250,19 @@ dataset_3 <-
   ) |>
   group_by(neighbourhood, collision_type) |>
   mutate(num = n()) |> # count number of rows based on neighbourhood and collision type
+  # Create a new column for the number of collisions for the neighbourhood for each 
+  # type of collision (2017 to 2023)
   rename(num_of_collisions = num) |>
   unique() |> # This will filter out repeated rows
   group_by(neighbourhood, year) |>
   mutate(num = n()) |> # count number of rows based on neighbourhood and year
+  # Create a new column for the yearly collisions number for the neighbourhood (2017 to 2023)
   rename(yearly_collision_num = num) |>
   unique() |>
-  group_by(neighbourhood) |>
-  mutate(num = n()) |>
-  rename(total_collisions_2017_2023 = num) |>
+  group_by(neighbourhood) |> # count number of rows based on neighbourhood
+  mutate(num = n()) |> 
+  # Create a new column for the total of collisions for the neighbourhood (2017 to 2023)
+  rename(total_collisions_2017_2023 = num) |> 
   unique()
 
 #### Save data ####

### 2024-01-22 scripts/03-test_data.R

@@ -1,15 +1,108 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
+# Purpose: Tests the datasets 
 # Author: Emily Su
 # Date: 23 January 2024
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: tidyverse has been installed by running 00-install_packages.R,
+# datasets have been downloaded by running 01-download_data.R, and datasets have 
+# been cleaned by running 02-data_cleaning.R
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
+#### Read in CSV files ####
+# Read in dataset 1 (Collisions Data)
+dataset_1 <- read_csv("outputs/data/cleaned_collisions_data.csv")
+# Read in dataset 2 (Ward Data)
+dataset_2 <- read_csv("outputs/data/cleaned_ward_data.csv")
+# Save dataset 3 (Map Data)
+dataset_3 <- read_csv("outputs/data/cleaned_map_data.csv")
 
 #### Test data ####
+
+# Tests on dataset 1 (Collision Data)
+# Expected Columns: year | collision_type | num_of_collisions 
+# 1. "year" does not contain years before 2017 and after 2023
+dataset_1$year |> min() == 2017
+dataset_1$year |> max() == 2023
+
+# 2. We have 4 types of collisions: "Personal Injury", "Fail to Remain", 
+# "Property Damage", "Fatal"
+dataset_1$collision_type |>
+  unique() |>
+  length() == 4
+
+dataset_1$collision_type |>
+  unique() %in% c("Personal Injury", "Fail to Remain", "Property Damage", "Fatal")
+
+# 3. num_of_collisions is greater than or equal to 0.
+dataset_1$num_of_collisions |> min() >= 0
+
+# 4. Check the classes
+class(dataset_1$year) == "numeric"
+class(dataset_1$num_of_collisions) == "numeric"
+class(dataset_1$collision_type) == "character"
+
+
+# Tests on dataset 2 (Ward Data)
+# Expected columns: ward_num | ward_name | pop_num | avg_income
+# 1. Smallest ward number is 1 and the largest ward number is 25
+dataset_2$ward_num |> min() == 1
+dataset_2$ward_num |> max() == 25
+
+# 2. There are 25 wards 
[…]

### 2024-01-22 scripts/02-data_cleaning.R
Updated script and csv filed to include pedestrian_involved

@@ -36,14 +36,16 @@ cleaned_collisions_data <-
     ftr_collisions, 
     long_wgs84,
     lat_wgs84, 
-    neighbourhood_158
+    neighbourhood_158,
+    pedestrian
   )
 
 # Renaming occ_year column to year
 cleaned_collisions_data <-
   cleaned_collisions_data |>
   rename( # Left side is the new name and the right side is the old name
-    year = occ_year
+    year = occ_year,
+    pedestrian_involved = pedestrian
   )
 
 # Get rid of all rows with NA 
@@ -65,18 +67,23 @@ cleaned_collisions_data$ftr_collisions <- str_replace(
 cleaned_collisions_data$ftr_collisions <- str_replace(
   cleaned_collisions_data$ftr_collisions, "NO", "0")
 
+# Replace NO and N\A with a 0 and YES with a 1 
+cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved != "YES"] <- 0
+cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved == "YES"] <- 1
+
 # Convert columns from character to numeric
 cleaned_collisions_data$injury_collisions <- as.numeric(cleaned_collisions_data$injury_collisions)
 cleaned_collisions_data$pd_collisions <- as.numeric(cleaned_collisions_data$pd_collisions)
 cleaned_collisions_data$ftr_collisions <- as.numeric(cleaned_collisions_data$ftr_collisions)
+cleaned_collisions_data$pedestrian_involved <- as.numeric(cleaned_collisions_data$pedestrian_involved)
 
 
 # Dataset 1 (number of collisions, collision types, and what year the collision occurred)
-# Expected Columns: year | collision_type | num_of_collisions 
+# Expected Columns: year | collision_type | num_of_collisions
 # Create a new dataset
 dataset_1 <-
   cleaned_collisions_data |>
-  select(-c(long_wgs84,lat_wgs84, neighbourhood_158)) |>
+  select(-c(long_wgs84,lat_wgs84, neighbourhood_158, pedestrian_involved)) |>
   group_by(year, fatalities, injury_collisions, pd_collisions, ftr_collisions) |>
   mutate(num = n()) |> # count number of rows based on premise and year
   rename(num_of_collisions = num) |>
@@ -215,7 +222,7 @@ cleaned_ward_data <-
 # Dataset 3 (Contains data of collision type, their longitude and latitude, 
[…]

### 2024-01-22 scripts/03-test_data.R

@@ -97,12 +97,21 @@ dataset_3$yearly_collision_num |> min() >= 0
 # 5. total_collisions_2017_2023 is greater than or equal to 0.
 dataset_3$total_collisions_2017_2023 |> min() >= 0
 
+# 6. pedestrian_involved only contains 0 and 1 
+dataset_3$pedestrian_involved |>
+  unique() %in% c(0, 1)
+
+dataset_3$pedestrian_involved |>
+  unique() |>
+  length() == 2
+
 # 6. Check classes 
 class(dataset_3$year) == "numeric"
 class(dataset_3$collision_type) == "character"
 class(dataset_3$long) == "numeric"
 class(dataset_3$lat) == "numeric"
 class(dataset_3$neighbourhood) == "character"
+class(dataset_3$pedestrian_involved) == "numeric"
 class(dataset_3$num_of_collisions) == "numeric"
 class(dataset_3$yearly_collision_num) == "numeric"
 class(dataset_3$total_collisions_2017_2023) == "numeric"

### 2024-01-22 scripts/02-data_cleaning.R
Updated scripts for cleaning and testig to account for class changes for pedestrian_involved

@@ -67,16 +67,17 @@ cleaned_collisions_data$ftr_collisions <- str_replace(
 cleaned_collisions_data$ftr_collisions <- str_replace(
   cleaned_collisions_data$ftr_collisions, "NO", "0")
 
-# Replace NO and N\A with a 0 and YES with a 1 
-cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved != "YES"] <- 0
-cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved == "YES"] <- 1
+# Replace N\A and "NO' with a "No" and "YES" with a "Yes"
+cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved != "YES"] <- "No"
+cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved != "YES"] <- "Yes"
 
 # Convert columns from character to numeric
 cleaned_collisions_data$injury_collisions <- as.numeric(cleaned_collisions_data$injury_collisions)
 cleaned_collisions_data$pd_collisions <- as.numeric(cleaned_collisions_data$pd_collisions)
 cleaned_collisions_data$ftr_collisions <- as.numeric(cleaned_collisions_data$ftr_collisions)
-cleaned_collisions_data$pedestrian_involved <- as.numeric(cleaned_collisions_data$pedestrian_involved)
 
+# Convert column pedestrian_involved from character to numeric
+cleaned_collisions_data$pedestrian_involved <- as.character(cleaned_collisions_data$pedestrian_involved)
 
 # Dataset 1 (number of collisions, collision types, and what year the collision occurred)
 # Expected Columns: year | collision_type | num_of_collisions

### 2024-01-22 scripts/03-test_data.R

@@ -111,7 +111,7 @@ class(dataset_3$collision_type) == "character"
 class(dataset_3$long) == "numeric"
 class(dataset_3$lat) == "numeric"
 class(dataset_3$neighbourhood) == "character"
-class(dataset_3$pedestrian_involved) == "numeric"
+class(dataset_3$pedestrian_involved) == "character"
 class(dataset_3$num_of_collisions) == "numeric"
 class(dataset_3$yearly_collision_num) == "numeric"
 class(dataset_3$total_collisions_2017_2023) == "numeric"

### 2024-01-22 scripts/02-data_cleaning.R
Fixed condition to replace column value for pedestrian_involved

@@ -69,7 +69,7 @@ cleaned_collisions_data$ftr_collisions <- str_replace(
 
 # Replace N\A and "NO' with a "No" and "YES" with a "Yes"
 cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved != "YES"] <- "No"
-cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved != "YES"] <- "Yes"
+cleaned_collisions_data$pedestrian_involved[cleaned_collisions_data$pedestrian_involved == "YES"] <- "Yes"
 
 # Convert columns from character to numeric
 cleaned_collisions_data$injury_collisions <- as.numeric(cleaned_collisions_data$injury_collisions)

### 2024-01-22 scripts/02-data_cleaning.R
removed unused code in data cleaning script

@@ -261,12 +261,10 @@ dataset_3 <-
   # Create a new column for the number of collisions for the neighbourhood for each 
   # type of collision (2017 to 2023)
   rename(num_of_collisions = num) |>
-  unique() |> # This will filter out repeated rows
   group_by(neighbourhood, year) |>
   mutate(num = n()) |> # count number of rows based on neighbourhood and year
   # Create a new column for the yearly collisions number for the neighbourhood (2017 to 2023)
   rename(yearly_collision_num = num) |>
-  unique() |>
   group_by(neighbourhood) |> # count number of rows based on neighbourhood
   mutate(num = n()) |> 
   # Create a new column for the total of collisions for the neighbourhood (2017 to 2023)

### 2024-01-23 scripts/02-data_cleaning.R
updated data cleaning script to clean city ward data

@@ -12,12 +12,17 @@
 library(tidyverse)
 library(janitor)
 library(dplyr)
+library(sf)
 
 #### Clean data ####
 
 ## Read in datasets ##
 raw_collisions_data <- read_csv("inputs/data/unedited_collisions_data.csv", show_col_types = FALSE)
 raw_ward_profiles_data <- read_csv("inputs/data/unedited_ward_profiles_data.csv", show_col_types = FALSE)
+raw_city_ward_data <- readRDS("inputs/data/city_wards.RDS")
+
+# Clean the column names for city ward data
+cleaned_city_ward_data <- clean_names(raw_city_ward_data)
 
 ## For collisions dataset ##
 # Clean the column names for collision data
@@ -219,12 +224,12 @@ cleaned_ward_data <-
   right_join(ward_avg_income, by = "ward_num") |>
   select(ward_num, ward_name, pop_num, avg_income) # Rearrange order
 
-
 # Dataset 3 (Contains data of collision type, their longitude and latitude, 
 # neighbourhood of collision, and number of collisions of each neighbourhood
 # from 2017 to 2023)
 # Expected: year | collision_type | long | lat | neighbourhood | pedestrian_involved
 # num_of_collisions | yearly_collision_num | total_collisions_2017_2023
+
 dataset_3 <- 
   cleaned_collisions_data |>
   unique() |> # This will filter out repeated rows  
@@ -249,9 +254,9 @@ dataset_3 <-
   select(-c(fatalities, injury_collisions, pd_collisions, ftr_collisions)) |> # Remove columns
   select(year, collision_type, long_wgs84, lat_wgs84, neighbourhood_158, pedestrian_involved) |> # Rearrange columns in desired order  
   rename ( # Left side is the new name and the right side is the old name
+    neighbourhood = neighbourhood_158,
     long = long_wgs84,
-    lat = lat_wgs84,
-    neighbourhood = neighbourhood_158
+    lat = lat_wgs84
   ) |>
   filter(
     (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
@@ -278,3 +283,5 @@ write_csv(dataset_1, "outputs/data/cleaned_collisions_data.csv")
 write_csv(cleaned_ward_data, "outputs/data/cleaned_ward_data.csv")
 # Save dataset 3
 write_csv(dataset_3, "outputs/data/cleaned_map_data.csv")
+# Save dataset 4
+saveRDS(cleaned_city_ward_data, "outputs/data/cleaned_city_wards.RDS")

### 2024-01-23 scripts/02-data_cleaning.R
Updated script and tests to included one more column for cleaned collision data

@@ -6,7 +6,7 @@
 # Contact: em.su@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: tidyverse has been installed by running 00-install_packages.R
-# and datasets have been downloaded by running 01-download_data.R. 
+# and datasets have been downloaded by running 01-download_data.R.
 
 #### Workspace setup ####
 library(tidyverse)
@@ -84,21 +84,25 @@ cleaned_collisions_data$ftr_collisions <- as.numeric(cleaned_collisions_data$ftr
 # Convert column pedestrian_involved from character to numeric
 cleaned_collisions_data$pedestrian_involved <- as.character(cleaned_collisions_data$pedestrian_involved)
 
-# Dataset 1 (number of collisions, collision types, and what year the collision occurred)
-# Expected Columns: year | collision_type | num_of_collisions
+# Dataset 1 (number of collisions, collision types, what year the collision occurred, 
+# and number of pedestrians involved in the crash)
+# Expected Columns: year | collision_type | num_of_collisions | num_of_pedestrians
 # Create a new dataset
 dataset_1 <-
   cleaned_collisions_data |>
-  select(-c(long_wgs84,lat_wgs84, neighbourhood_158, pedestrian_involved)) |>
+  select(-c(long_wgs84,lat_wgs84, neighbourhood_158)) |>
   group_by(year, fatalities, injury_collisions, pd_collisions, ftr_collisions) |>
   mutate(num = n()) |> # count number of rows based on premise and year
   rename(num_of_collisions = num) |>
-  unique() |> # This will filter out repeated rows  
+  mutate(
+    num_of_pedestrians = sum(pedestrian_involved == "Yes")
+  ) |>
+  unique() |> # This will filter out repeated rows
   filter(
-    (fatalities == 1 && injury_collisions == 0 && pd_collisions == 0 && ftr_collisions == 0) |
-      (fatalities == 0 && injury_collisions == 1 && pd_collisions == 0 && ftr_collisions == 0) |
-      (fatalities == 0 && injury_collisions == 0 && pd_collisions == 1 && ftr_collisions == 0) |
-      (fatalities == 0 && injury_collisions == 0 && pd_collisions == 0 && ftr_collisions == 1)
+    (fatalities == 1 & injury_collisions == 0 & pd_collisions == 0 & ftr_collisions == 0) |
[…]

### 2024-01-23 scripts/03-test_data.R

@@ -23,7 +23,7 @@ dataset_3 <- read_csv("outputs/data/cleaned_map_data.csv")
 #### Test data ####
 
 # Tests on dataset 1 (Collision Data)
-# Expected Columns: year | collision_type | num_of_collisions 
+# Expected Columns: year | collision_type | num_of_collisions | num_of_pedestrians
 # 1. "year" does not contain years before 2017 and after 2023
 dataset_1$year |> min() == 2017
 dataset_1$year |> max() == 2023
@@ -40,10 +40,15 @@ dataset_1$collision_type |>
 # 3. num_of_collisions is greater than or equal to 0.
 dataset_1$num_of_collisions |> min() >= 0
 
-# 4. Check the classes
+# 3. num_of_pedestrians is greater than or equal to 0.
+dataset_1$num_of_pedestrians |> min() >= 0
+
+# 5. Check the classes
 class(dataset_1$year) == "numeric"
 class(dataset_1$num_of_collisions) == "numeric"
 class(dataset_1$collision_type) == "character"
+class(dataset_1$num_of_pedestrians) == "integer"
+
 
 
 # Tests on dataset 2 (Ward Data)

### 2024-01-23 outputs/paper/paper.qmd
Uploaded current progress on Paper and updated script comments

--- 2024-01-19 outputs/paper/paper.qmd
@@
 ---
-title: "(Working Title) Motor Vehicle Collision Across Toronto"
-subtitle: "My subtitle if needed"
+title: "Motor Vehicle Collisions Reduced Across Toronto During and After the Beginning of the COVID-19 Pandemic"
 author: "Emily Su"
@@
 date-format: "D MMMM YYYY"
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestarian safety are growing concerns in Toronto espically after the COVID-19 Pandemic.
+This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.
+The results shows that motor vehicle collisions and pedestarian involvement in them have decreased after the pandemic.
+However, further investigation is needed on the demographics of Toronto areas with high number of motor vehicle collisions."
 format:
@@
 ---
+```{r}
+#| include: false
+#| warning: false
+#| message: false
+#### Preamble ####
+# Purpose: To create a graphs of the number of collisions and the types of collisions
+# in Toronto in different neighbourhoods from 2017 to 2023.
+# Author: Emily Su
+# Email: em.su@mail.utoronto.ca
+# Date: 23 January 2024
+# Prerequisites: All the scripts in the scripts folder have been ran starting from 00-install_packages.R.
+```
 
@@
 
-library(tidyverse)
-library(palmerpenguins)
+library("tidyverse") # Contains data-related packages
+library("knitr") # To make tables
+library("opendatatoronto")
+library("janitor") # To clean datasets
+library("dplyr")
+library("ggplot2") # To make graphs
+library("sf") # To make maps
 ```
@@
 
-You can and should cross-reference sections and sub-sections.
+As of 2024, Toronto was reported to have one of the worst traffic congestion in North America [@citeWorstCongestion].
+As Toronto's population increased over the years, congestion on the road increased and road and pedestrian safety is a growing concern for many.
+Some also believe that roads have gotten unsafe after the pandemic.
+In order to reduce major injuries and traffic collision fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating fatalities on roads all together [@visionzero].
+An ambitious goal however it raises the following question: how does motor vehicle collisions (MVC) compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?
 
-The remainder of this paper is structured as follows. @sec-data....
+In this paper, I analyzed collisions trends from 2017 to 2023 specifically into collisions in Toronto neighbourhoods and wards, the types of collisions that occurred, and the number of pedestrians involved.
+In my findings, from 2020 onwards, there was a decline in collisions...
+
+By understanding collision trends across Toronto and over the years, it can help improved safety plans for strategies like Vision Zero for instance for pedestrians in specific areas of Toronto.
 
 # Data {#sec-data}
+The data used by the paper are `City Wards` [@citecitywards], `Ward Profiles (25-Ward Model)` [@citewardprofiles], and `Police Annual Statistical Report - Traffic Collisions` [@citecollisionsdata]  datasets, which were sourced from Open Data Toronto Portal.
+These datasets were accessed using the `opendatatoronto` library [@citeopendatatoronto].
+Another dataset that could have been used but was not was the `About Motor Vehicle Collisions involving Killed or Seriously Injured Persons` dataset from Open Data Toronto Portal.
+However, this dataset's most current record was in 2022 and is outdated for our purposes.
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+The datasets were cleaned and analyzed using `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
+Some of the datasets used for the data analysis combined the `Ward Profiles (25-Ward Model)` and `Police Annual Statistical Report - Traffic Collisions` in order to create datasets with the following variables: `collision_type`, `num_of_collisions`, `num_of_pedestrians`, `yearly_collision_num`, and `total_collisions_2017_2023`.
+`collision`.
+These variables were formed by mostly created group existing columns and counting the number of rows with certain column variables.
+The Toronto Police Services has defined MVC under 4 different categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata].
+Fatal means that a person was killed during the vehicle collision, personal injury indicates a person was injured during the collision, fail to remain means that the person left the scene before after information was provided regarding the collision, and property damage
+indicates that property has been damaged during the collision.
+With this, the 4 types are possible values under the variable `collision_type`.
+`num_of_collisions` represents the number of collisions that occurred in either a neighbhourhood in a specific year or the number of collisions of a certain type of a certain year depending on the context.
+`num_of_pedestrians` represents the number of pedestrians that were hit in the vehicle collision.`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood mentioned in the `Police Annual Statistical Report - Traffic Collisions` dataset.
+For the dataset regarding Toronto ward profiles, the following variables were created: `ward_name` (indicates the name of the Toronto ward), `pop_num` (represents the population size of Toronto wards as of 2020), and `avg_income` (signifies the average household income of each ward as of 2020).
+
+## Results
+```{r}
+#| include: false # Would run the code but it will have no output and code is not in the document 
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included
+
+# Read in data
+
+# Map Data
+the_map <- readRDS('../data/cleaned_city_wards.RDS')
+og_cleaned_map_data <- read_csv('../data/cleaned_map_data.csv', show_col_types = FALSE)
+cleaned_map_data <-
+  og_cleaned_map_data |>
+  filter(
+    (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
+  ) 
+cleaned_map_data <- st_as_sf(cleaned_map_data, coords = c('long', 'lat'), crs = 4326)
+
+# Collision Data 
+cleaned_collisions_data <- read_csv("../data/cleaned_collisions_data.csv", show_col_types = FALSE)
+
+# Ward Data
+cleaned_ward_data <- read_csv("../data/cleaned_ward_data.csv", show_col_types = FALSE)
+```
+
+### Collisions across Toronto neighbourhoods and Wards (2017-2023) {#sec-location}
+There are 25 wards and 158 neighbourhoods in the city of Toronto. The distribution of MVC across different neighbourhoods and wards was investigated and the following was obtained:
+```{r}
+#| label: tbl-neighbhourhoodtable
+#| tbl-cap: Toronto Neighbourhoods with the most number of motor vehicle collisions from 2017 to 2023
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included 
+# To create table corresponding to the map
+
+collisions_num_per_neighbhourhood <- 
+arrange(og_cleaned_map_data, desc(total_collisions_2017_2023)) |>
+  select(neighbourhood, total_collisions_2017_2023) |>
+  unique() |>
+  rename (
+    "Neighbourhood" = neighbourhood,
+    "Number of Motor Vehicle Collisions (2017 to 2023)" = total_collisions_2017_2023
+  )
+
+collisions_num_per_neighbhourhood |>
+  head() |>
+  kable() 
+```
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| label: fig-numcollisionmap
+#| fig-cap: This figure shows seperate maps of the number of motor vehicle collisions from 2017 to 2023 across different Toronto wards. 
   #| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
+  #| warning: false
+  #| message: false
+  # Code snippet to add points to map is from https://stackoverflow.com/questions/58676661/geom-sf-mapping-points-shapes
+  # Code snippet to change colour of plots 
+  # https://www.rdocumentation.org/packages/ggplot2/versions/2.2.1/topics/scale_colour_gradient
+# Code snippet to change theme adapated from  
+# https://ggplot2.tidyverse.org/reference/theme.html
+ggplot() + 
+  geom_point(data = cleaned_map_data,
+    aes(color = num_of_collisions, geometry = geometry), alpha = 0.05,
+    stat = "sf_coordinates"
+  ) +
+  facet_wrap(vars(year), ncol = 4) +
+  # Change colours of the points
+  scale_color_gradient(low = 'lightblue', high = 'darkblue', name = "Number of Collisions") +
+  # Overlays a map of Toronto  
+  geom_sf(data = the_map, fill = "transparent") +
+  # Add ward number labels 
+  geom_sf_text(data=the_map, aes(label = area_short_code), color = '#333333', size = 1.5) + 
+  # Changes the font of the title and removes the x and y axis 
+  theme(plot.title = element_text(size = 9, face = 'bold'), rect = element_blank(), axis.text.x =
+        element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.title = element_text(size = 7)) + 
+  labs(x = "Longitude",
+       y = "Latitude", 
+       title="Number of Motor Vehicle Collisions Across Toronto Wards from 2017 to 2023") +
+  theme(legend.position = c(0.9, 0.2))
 ```
 
-Talk more about it.
+```{r}
+#| label: tbl-wardprofilestats
+#| tbl-cap: Summary Statistics of Toronto Ward Profile in 2020 
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+cleaned_ward_data |>
+  select(pop_num, avg_income) |>
+  rename (
+    "Population Number" = pop_num, 
+    "Average Household Income"= avg_income
+  ) |>
+  summary() |>
+  kable()
+```
+In @tbl-neighbhourhoodtable, West Humber-Clairville has the greatest number of collisions with 2812 MVC from 2017 to 2023.
+West Humber-Clairville is located in ward 1, which from @fig-numcollisionmap is the region with the darkests shade of blue and is the most dense in terms of number of collisions [@citefinder].
+Following this is Wexford/Maryvale, which is located in ward 21 with 2075 collisions.
+From @tbl-wardprofile, something interesting to note is that ward 1's average household income is \$38,135 with a population of 115120 people and ward 21's average household income is \$40,565 with a population of 115120 people.
+
+### Type of Collisions in Toronto (2017 - 2023) {#sec-types}
+```{r}
+#| label: tbl-collisiontype
+#| tbl-cap: Number of Motor Vehicle Collisions of Different Types Across Toronto from 2017 to 2023
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included
+
+cleaned_collisions_data_table <-
+cleaned_collisions_data |>
+  group_by(year) |>
+  unique() |>
+  mutate(
+    proportion =  round(num_of_collisions / sum(num_of_collisions), 2)
+  ) |>
+  ungroup() |>
+  rename (
+    "Year" = year,
+    "Collision Type" = collision_type,
+    "Number of Collisions" = num_of_collisions,
+    "Proportion of Collisions (Per Year)" = proportion
+  ) |>
+  select("Year", "Collision Type", "Number of Collisions", "Proportion of Collisions (Per Year)")
+cleaned_collisions_data_table |>
+  kable()
+```
 
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: tbl-collisionstats
+#| tbl-cap: Summary Statistics of the Number of Motor Vehicle Collisions of Different Types Across Toronto from 2017 to 2023
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included 
+cleaned_collisions_data_table |>
+  select("Number of Collisions") |>
+  summary() |>
+  kable()
+```
+
+```{r}
+#| label: fig-collisionhistogram
+#| fig-cap: This histogram reveals the number of motor vehicle collisions from 2017 to 2023 across Toronto and the different types of motor vehicle collisions. 
+#| echo: false
+#| warning: false
+#| message: false
+# Referenced chapter Chapter 5 of Telling Stories with Data by Rohan Alexander for 
+# scale_fill_viridis_d(), theme(), mapping = aes(...)
+cleaned_collisions_data |>
+  ggplot(mapping = aes(x = year, y = num_of_collisions, fill = collision_type)) +
+  geom_bar(stat="identity") +
+  ggtitle("Number of Motor Vehicle Collisions in Toronto from 2017 to 2023") +
+  labs(x = "Year", y = "Number of Collisions", fill = "Type of Collisions") +
+  theme(legend.position = "bottom", plot.title = element_text(size = 9, face = "bold"), rect = element_blank(), legend.title = element_text(size = 9), legend.text = element_text(size = 7)) +
+  scale_fill_viridis_d()
+```
+From @fig-collisionhistogram, when comparing the overall number of collisions before 2020 and 2020 onwards, there is a overall annual decrease with the overall number of collisions.
+However, collisions involving property damage remains the majority of collision cases over the years.
+We can see this decrease from 2019 and 2020 from about 79205 collisions to 43297 collisions.
+According to @tbl-collisionstats and @tbl-collisiontype, the peak number of property damage collisions was at 60865 in 2019, which made up 77% of the collisions in that year.
+However property damage collisions in 2020 was up to 79% of collisions in 2020.
+
+### Pedestrians Involved in a Motor Vehicle Collision (2017-2023) {#sec-pedes}
+
+```{r}
+#| label: tbl-pedestrianhistogram
+#| tbl-cap: Number of Pedestrians Involved in a Motor Vehicle Collision from 2017 to 2023
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included
+
+# Add up num of pedestrians for each year, get number of collisions each year, and the proportion
+cleaned_collisions_data_yearly <-
+  cleaned_collisions_data |>
+  group_by(year) |>
+  unique() |>
+  mutate (
+    yearly_num_pedestrians = sum(num_of_pedestrians)
+  ) |>
+  mutate(
+    num_of_collisons_year = sum(num_of_collisions)
+  ) |>
+  mutate(
+    proportion = round(yearly_num_pedestrians / num_of_collisons_year, 3)
+  ) |>
+  ungroup()
+
+cleaned_collisions_data_yearly |>
+  select(year, yearly_num_pedestrians, proportion) |>
+  unique() |>
+  rename(
+    "Year" = year,
+    "Number of pedestrians involved (Per Year)" = yearly_num_pedestrians, 
+    "Proportion of pedestrians involved (Per Year)" = proportion
+  ) |>
+  kable()
+```
+
+```{r}
+#| label: fig-pedestrianhistogram
+#| fig-cap: This histogram reveals the number of pedestrians involved in a motor vehicle collisions from 2017 to 2023 across Toronto. 
 #| echo: false
@@
 
-analysis_data <- read_csv(here::here("outputs/data/analysis_data.csv"))
+# Referenced tutorial notes from summer STA238 2023 to obtain the code portion of geom_bar(stat="identity") and plot.title
+# Referenced chapter Chapter 5 of Telling Stories with Data by Rohan Alexander for 
+# scale_fill_viridis_d(), theme(), mapping = aes(...)
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
-  theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+cleaned_collisions_data_yearly |>
+  ggplot(mapping = aes(x = year, y = yearly_num_pedestrians)) +
+  geom_col(fill="lightblue",colour="lightblue") +
+  ggtitle("Pedestrians involved in Motor Vehicle Collisions in Toronto from 2017 to 2023") +
+  labs(x = "Year", y = "Number of Collisions Involved") +
+  theme(plot.title = element_text(size = 9, face = "bold"), rect = element_blank(), legend.title = element_text(size = 9), legend.text = element_text(size = 7)) 
 ```
 
-Talk way more about it.
-
-# Model
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("outputs/models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
+\newpage
 
@@
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+According to @tbl-wardprofilestats and @sec-location, ward 1 and ward 21's population number is above the mean of 110452 people (also known as average) however both wards are below the mean of $46,436 in terms of average household income.
 
@@
 
-Weaknesses and next steps should also be included.
-
 \newpage
@@
 # Appendix {-}
+```{r}
+#| label: tbl-wardprofile
+#| tbl-cap: Toronto Ward Profile in 2020
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included
 
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
+cleaned_ward_data |>
+  rename (
+    "Ward Number" = ward_num,
+    "Ward Name" = ward_name,
+    "Population Number" = pop_num, 
+    "Average Household Income"= avg_income
+  ) |>
+  kable()
 ```

### 2024-01-23 outputs/paper/paper.qmd
Updated paper, references, and README.md

@@
 date-format: "D MMMM YYYY"
-abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestarian safety are growing concerns in Toronto espically after the COVID-19 Pandemic.
+abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestarian safety are growing concerns in Toronto, espically after the COVID-19 Pandemic.
 This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.

### 2024-01-23 scripts/02-data_cleaning.R

@@ -85,7 +85,7 @@ cleaned_collisions_data$ftr_collisions <- as.numeric(cleaned_collisions_data$ftr
 cleaned_collisions_data$pedestrian_involved <- as.character(cleaned_collisions_data$pedestrian_involved)
 
 # Dataset 1 (number of collisions, collision types, what year the collision occurred, 
-# and number of pedestrians involved in the crash)
+# and number of pedestrians involved in the collision)
 # Expected Columns: year | collision_type | num_of_collisions | num_of_pedestrians
 # Create a new dataset
 dataset_1 <-

### 2024-01-23 outputs/paper/paper.qmd
Fixed spelling in current paper and fixed README.md formatting

--- 2024-01-23 outputs/paper/paper.qmd
@@
 date-format: "D MMMM YYYY"
-abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestarian safety are growing concerns in Toronto espically after the COVID-19 Pandemic.
+abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestarian safety are growing concerns in Toronto, especially after the COVID-19 Pandemic.
 This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.
-The results shows that motor vehicle collisions and pedestarian involvement in them have decreased after the pandemic.
-However, further investigation is needed on the demographics of Toronto areas with high number of motor vehicle collisions."
+The results shows that motor vehicle collisions and pedestrian involvement in them have decreased after the pandemic.
+However, further investigations is needed on the demographics of Toronto areas with high number of motor vehicle collisions."
 format:

### 2024-01-23 outputs/paper/paper.qmd
Updated wording in introduction of paper

@@
 
-By understanding collision trends across Toronto and over the years, it can help improved safety plans for strategies like Vision Zero for instance for pedestrians in specific areas of Toronto.
+By understanding collision trends across Toronto and over the years, it can assist with how current safety plans like Vision Zero can improve and assist with being more strategic such as safety plans in specific locations of Toronto.
 

### 2024-01-23 outputs/paper/paper.qmd
Fixed grammar in discussion

@@
 
-According to @tbl-wardprofilestats and @sec-location, ward 1 and ward 21's population number is above the mean of 110452 people (also known as average) however both wards are below the mean of $46,436 in terms of average household income.
+According to @tbl-wardprofilestats and @sec-location, ward 1 and ward 21's population number are above the mean of 110452 people (also known as average) however both wards are below the mean of $46,436 in terms of average household income.
 

### 2024-01-23 outputs/paper/paper.qmd
Added R citation to paper

@@
 
-The datasets were cleaned and analyzed using `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
+The datasets were cleaned and analyzed using the statistical programming language `R` [@citeR], `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
 Some of the datasets used for the data analysis combined the `Ward Profiles (25-Ward Model)` and `Police Annual Statistical Report - Traffic Collisions` in order to create datasets with the following variables: `collision_type`, `num_of_collisions`, `num_of_pedestrians`, `yearly_collision_num`, and `total_collisions_2017_2023`.

### 2024-01-23 outputs/paper/paper.qmd
Updated README.md to remove authorship and update data section

--- 2024-01-23 outputs/paper/paper.qmd
@@
 
-The datasets were cleaned and analyzed using `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
+The datasets were simulated, cleaned, analyzed, and tested using the statistical programming language `R` [@citeR], `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
 Some of the datasets used for the data analysis combined the `Ward Profiles (25-Ward Model)` and `Police Annual Statistical Report - Traffic Collisions` in order to create datasets with the following variables: `collision_type`, `num_of_collisions`, `num_of_pedestrians`, `yearly_collision_num`, and `total_collisions_2017_2023`.

### 2024-01-24 scripts/02-data_cleaning.R
Added Alexander (2023) to reference list and updated README.md's repo structure

@@ -181,6 +181,8 @@ ward_avg_income_row <-
 # Convert row to a column 
 ward_avg_income <- 
   ward_avg_income_row |>
+  # Code snippet to convert to a 1-column matrix and then to a vector 
+  # was adapted from the following: https://www.geeksforgeeks.org/convert-matrix-to-vector-in-r/
   t() |> # Convert 1-row matrix into a 1-column matrix
   c() |> # Convert to a vector
   tibble() |> # Change into a tibble

### 2024-01-24 scripts/02-data_cleaning.R
Added in-text citation in code for Alexander (2023)

@@ -16,6 +16,8 @@ library(sf)
 
 #### Clean data ####
 
+# Code below referenced Alexander (2023) 
+
 ## Read in datasets ##
 raw_collisions_data <- read_csv("inputs/data/unedited_collisions_data.csv", show_col_types = FALSE)
 raw_ward_profiles_data <- read_csv("inputs/data/unedited_ward_profiles_data.csv", show_col_types = FALSE)

### 2024-01-25 outputs/paper/paper.qmd
Updated paper with current progress

--- 2024-01-23 outputs/paper/paper.qmd
@@
 ---
-title: "Motor Vehicle Collisions Reduced Across Toronto During and After the Beginning of the COVID-19 Pandemic"
+title: "Motor Vehicle Collisions Decreased in Toronto During and After the Beginning of the COVID-19 Pandemic"
 author: "Emily Su"
@@
 date-format: "D MMMM YYYY"
-abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestarian safety are growing concerns in Toronto, especially after the COVID-19 Pandemic.
+abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestrian safety are growing concerns in Toronto, especially after the COVID-19 Pandemic.
 This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.
-The results shows that motor vehicle collisions and pedestrian involvement in them have decreased after the pandemic.
-However, further investigations is needed on the demographics of Toronto areas with high number of motor vehicle collisions."
+The results shows that motor vehicle collisions and pedestrian involvement in them have decreased after the pandemic but are still prevalent in more lower-income areas.
+Further investigation is needed on the demographics of Toronto areas with high number of motor vehicle collisions."
 format:
@@
 #### Preamble ####
-# Purpose: To create a graphs of the number of collisions and the types of collisions
-# in Toronto in different neighbourhoods from 2017 to 2023.
+# Purpose: To create graphs and analyze the number of collisions in different Toronto locations and the types of collisions rom 2017 to 2023.
 # Author: Emily Su
@@
 
-# Introduction
-
-As of 2024, Toronto was reported to have one of the worst traffic congestion in North America [@citeWorstCongestion].
-As Toronto's population increased over the years, congestion on the road increased and road and pedestrian safety is a growing concern for many.
-Some also believe that roads have gotten unsafe after the pandemic.
-In order to reduce major injuries and traffic collision fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating fatalities on roads all together [@visionzero].
-An ambitious goal however it raises the following question: how does motor vehicle collisions (MVC) compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?
-
-In this paper, I analyzed collisions trends from 2017 to 2023 specifically into collisions in Toronto neighbourhoods and wards, the types of collisions that occurred, and the number of pedestrians involved.
-In my findings, from 2020 onwards, there was a decline in collisions...
-
-By understanding collision trends across Toronto and over the years, it can assist with how current safety plans like Vision Zero can improve and assist with being more strategic such as safety plans in specific locations of Toronto.
-
-# Data {#sec-data}
-The data used by the paper are `City Wards` [@citecitywards], `Ward Profiles (25-Ward Model)` [@citewardprofiles], and `Police Annual Statistical Report - Traffic Collisions` [@citecollisionsdata]  datasets, which were sourced from Open Data Toronto Portal.
-These datasets were accessed using the `opendatatoronto` library [@citeopendatatoronto].
-Another dataset that could have been used but was not was the `About Motor Vehicle Collisions involving Killed or Seriously Injured Persons` dataset from Open Data Toronto Portal.
-However, this dataset's most current record was in 2022 and is outdated for our purposes.
-
-The datasets were cleaned and analyzed using `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
-Some of the datasets used for the data analysis combined the `Ward Profiles (25-Ward Model)` and `Police Annual Statistical Report - Traffic Collisions` in order to create datasets with the following variables: `collision_type`, `num_of_collisions`, `num_of_pedestrians`, `yearly_collision_num`, and `total_collisions_2017_2023`.
-`collision`.
-These variables were formed by mostly created group existing columns and counting the number of rows with certain column variables.
-The Toronto Police Services has defined MVC under 4 different categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata].
-Fatal means that a person was killed during the vehicle collision, personal injury indicates a person was injured during the collision, fail to remain means that the person left the scene before after information was provided regarding the collision, and property damage
-indicates that property has been damaged during the collision.
-With this, the 4 types are possible values under the variable `collision_type`.
-`num_of_collisions` represents the number of collisions that occurred in either a neighbhourhood in a specific year or the number of collisions of a certain type of a certain year depending on the context.
-`num_of_pedestrians` represents the number of pedestrians that were hit in the vehicle collision.`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood mentioned in the `Police Annual Statistical Report - Traffic Collisions` dataset.
-For the dataset regarding Toronto ward profiles, the following variables were created: `ward_name` (indicates the name of the Toronto ward), `pop_num` (represents the population size of Toronto wards as of 2020), and `avg_income` (signifies the average household income of each ward as of 2020).
-
-## Results
 ```{r}
@@
   filter(
-    (round(long) != 0 & round(lat) != 0) # Filter out long and lat that are at 0 
+    (round(long) != 0 & round(lat) != 0) # Filter out longitude and latitude that are at 0 
   ) 
@@
 
-### Collisions across Toronto neighbourhoods and Wards (2017-2023) {#sec-location}
-There are 25 wards and 158 neighbourhoods in the city of Toronto. The distribution of MVC across different neighbourhoods and wards was investigated and the following was obtained:
+# Introduction
+
+As of 2024, Toronto was reported to have one of the worst traffic congestion in North America [@citeWorstCongestion].
+As Toronto's population increases over the years, congestion on the road increased and road and pedestrian safety is a growing concern for many.
+Some also believe that roads have gotten unsafer in the years after the pandemic.
+In a 2023 study done by Ipsos surveying Canadians on road safety, they mentioned that concerns about road safety has increased to 65% in 2023 from 59% in 2021 [@citeIpsos].
+In order to reduce traffic collision injuries and fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating fatalities on roads all together [@visionzero].
+An ambitious goal yet it raises the question how effective is this plan across Toronto over the years.
+We investigated this question and the following: how does motor vehicle collisions compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?
+
+In this paper, collisions trends were analyzed from 2017 to 2023 specifically into motor vehicle collisions in Toronto neighbourhoods and wards, the types of motor vehicle collisions that occurred, and the number of pedestrians involved in them.
+In our findings, there was a decline in motor vehicle collisions overall from 2020 onwards.
+However, we saw that the locations motor vehicle collisions occurred at stayed consistent over the years.
+We also found that the wards that make up a greater proportion of motor vehicle collisions had average household incomes that are below Toronto's overall average household income.
+By understanding collision trends over the years in Toronto, it can aid with identifying areas of improvement of current safety plans like Vision Zero and overall helping others feel safer in the city.
+
+In the rest of this paper, the data section will cover the datasets used, define variables used by our tables and graphs, and briefly explain the data cleaning process.
+In the results section, we will reveal tables and graphs made on our datasets and explain what they show.
+In the discussion section, we will connect back to the real world and explain what the results could mean, the implication of our results, potential areas of improvement for the paper, and suggestions for future works.
+Finally, the appendix section will include the entire table of some of our results.
+
+# Data {#sec-data}
+The data used by the paper are `City Wards` [@citecitywards], `Ward Profiles (25-Ward Model)` [@citewardprofiles], and `Police Annual Statistical Report - Traffic Collisions` [@citecollisionsdata]  datasets, which were sourced from the City of Toronto's Open Data Portal.
+These datasets were accessed using the `opendatatoronto` library [@citeopendatatoronto].
+Another dataset that could have been used was the `About Motor Vehicle Collisions involving Killed or Seriously Injured Persons` dataset from Open Data Toronto Portal.
+However, this dataset's most current record was in 2022 and is outdated for our purposes.
+
+The datasets were simulated, cleaned, analyzed, and tested using the `R` programming language [@citeR], `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
+
+Some of the datasets used for the data analysis combined the `Ward Profiles (25-Ward Model)` and `Police Annual Statistical Report - Traffic Collisions` in order to create datasets with the following variables: `collision_type`, `num_of_collisions`, `num_of_pedestrians`, `yearly_collision_num`, and `total_collisions_2017_2023`.
+`collision`.
+These variables were formed by mostly created group existing columns and counting the number of rows with certain column variables.
+The Toronto Police Services has defined motor vehicle collisions under 4 different categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata].
+Fatal means that a person was killed during the vehicle collision, personal injury indicates a person was injured during the collision, fail to remain means that the person left the scene before after information was provided regarding the collision, and property damage indicates that property has been damaged during the collision.
+With this, the 4 types are possible values under the variable `collision_type`.
+`num_of_collisions` represents the number of collisions that occurred in either a neighbhourhood in a specific year or the number of collisions of a certain type of a certain year depending on the context.
+Due to the complexity of motor vehicle collisions and the different combinations of ways they can be categorized, our analysis looked at collisions that were defined in only 1 of the 4 categories.
+
+`num_of_pedestrians` represents the number of pedestrians that were hit in the vehicle collision.`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood mentioned in the `Police Annual Statistical Report - Traffic Collisions` dataset.
+For the dataset regarding Toronto ward profiles, the following variables were created: `ward_name` (indicates the name of the Toronto ward), `pop_num` (represents the population size of Toronto wards as of 2020), and `avg_income` (signifies the average household income of each ward as of 2020).
+
+# Results {#sec-result}
+### Motor vehicle collisions across Toronto neighbourhoods and wards (2017-2023) {#sec-location}
+As of 2024, There are 25 wards and 158 neighbourhoods in the city of Toronto.
+The distribution of motor vehicle collisions across different neighbourhoods and wards was investigated and the following was obtained:
 ```{r}
 #| label: tbl-neighbhourhoodtable
-#| tbl-cap: Toronto Neighbourhoods with the most number of motor vehicle collisions from 2017 to 2023
+#| tbl-cap: Toronto neighbourhoods with the most number of motor vehicle collisions from 2017 to 2023
 #| echo: false # Code will run and its output will show but the code itself will not be included in the doc
@@
 collisions_num_per_neighbhourhood <- 
+  # Order the data based on the total number of collisions from 2017 to 2023
   arrange(og_cleaned_map_data, desc(total_collisions_2017_2023)) |>
+  # Choose the columns neighbourhood and total_collisions_2017_2023 to include. 
   select(neighbourhood, total_collisions_2017_2023) |>
+  # Keep unique rows 
   unique() |>
-  rename (
-    "Neighbourhood" = neighbourhood,
-    "Number of Motor Vehicle Collisions (2017 to 2023)" = total_collisions_2017_2023
+  # Create new column for the proportion of motor vehicle collisions 
+  mutate( 
+    proportion =  round(total_collisions_2017_2023 / sum(total_collisions_2017_2023), 2)
+  ) |>
+  rename(
+    "Neighbourhood Name and Number" = neighbourhood,
+    "Number of Motor Vehicle Collisions (2017 to 2023)" = total_collisions_2017_2023,
+    "Proportion of Motor Vehicle Collisions (2017 to 2023)"  = proportion
   )
@@
 ```{r}
+#| label: tbl-wardprofilestats
+#| tbl-cap: Minimum, quartiles, median, and maximum of population size and average household income across Toronto wards in 2020 
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included
+
+cleaned_ward_data |>
+  # Choose pop_num and avg_income as columns to show 
+  select(pop_num, avg_income) |>
+  rename (
+    "Population Number" = pop_num, 
+    "Average Household Income (CAD)"= avg_income
+  ) |>
+  summary() |>
+  kable()
+```
+```{r}
 #| label: fig-numcollisionmap
-#| fig-cap: This figure shows seperate maps of the number of motor vehicle collisions from 2017 to 2023 across different Toronto wards. 
+#| fig-cap: Number of motor vehicle collisions from 2017 to 2023 across different Toronto wards. 
 #| echo: false
@@
 ggplot() + 
+  # Adds in points regarding number of collisions to graph 
   geom_point(data = cleaned_map_data,
@@
   ) +
+  # Creates different graphs for each year 
   facet_wrap(vars(year), ncol = 4) +
@@
   theme(plot.title = element_text(size = 9, face = 'bold'), rect = element_blank(), axis.text.x =
-        element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.title = element_text(size = 7)) + 
+        element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.title = 
+          element_text(size = 7)) + 
   labs(x = "Longitude",
@@
 ```
+@tbl-neighbhourhoodtable reveals that the neighbourhood with the highest number of motor vehicle collisions and the approximately how many motor vehicle collisions from 2017 to 2023 occurred in each neighbourhood.
+The neighbourhoods with a high number of collisions are the following: West Humber-Clairville (approximately 2%), Wexford/Maryvale (approximately 2%), South Riverdale (approximately 2%), York University Heights (approximately 2%), Clairlea-Birchmount (approximately 2%), Birchmount (approximately 1%), and Etobicoke City Centre (approximately 1%).
+The full table of @tbl-neighbhourhoodtable can be found here @tbl-neighbhourhoodtablefull in the Appendix section with the rest of the neighbourhoods making up approximately 1% of collisions or less.
 
-```{r}
-#| label: tbl-wardprofilestats
-#| tbl-cap: Summary Statistics of Toronto Ward Profile in 2020 
-#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
-#| warning: false # Warnings are not included 
-#| message: false # Messages are not included
+In @tbl-neighbhourhoodtable, West Humber-Clairville has the largest amount of collisions with 2812 motor vehicle collisions from 2017 to 2023.
+West Humber-Clairville is located in ward 1, which from @fig-numcollisionmap is the region with the darkests shade of blue and has the highest density of points [@citefinder].
+Following this is Wexford/Maryvale, which is located in ward 21 with 2075 collisions [@citefinder].
+The rest of the neighbourhood are in the following wards based on @citefinder: South Riverdale is in Ward 14, York University Heights is in Ward 7, Clairlea-Birchmount is in Ward 20, and Etobicoke City Centre is in Ward 2.
 
-cleaned_ward_data |>
-  select(pop_num, avg_income) |>
-  rename (
-    "Population Number" = pop_num, 
-    "Average Household Income"= avg_income
-  ) |>
-  summary() |>
-  kable()
-```
-In @tbl-neighbhourhoodtable, West Humber-Clairville has the greatest number of collisions with 2812 MVC from 2017 to 2023.
-West Humber-Clairville is located in ward 1, which from @fig-numcollisionmap is the region with the darkests shade of blue and is the most dense in terms of number of collisions [@citefinder].
-Following this is Wexford/Maryvale, which is located in ward 21 with 2075 collisions.
-From @tbl-wardprofile, something interesting to note is that ward 1's average household income is \$38,135 with a population of 115120 people and ward 21's average household income is \$40,565 with a population of 115120 people.
+@tbl-wardprofilestats reveals that the mean (also known as the average) population size per ward in Toronto is 110452 people with a standard deviation of 10594 people and the mean for the average household income per ward is \$46,436 with a standard deviation of \$11,647.51.
+It also indicates that the median for the population size per ward is 110095 people and the median for average household income per ward is $45,345.
+From @tbl-wardprofile, ward 1's average household income is \$38,135 with a population of 115120 people and ward 21's average household income is \$40,565 with a population of 115120 people.
+For ward 14, 7, 20, and 2, the average households incomes are \$46,235, \$37,675, \$41,905, and \$45,345, respectively and the population sizes are 104555, 111200, 110095, and 117200 people, respectively.
 
-### Type of Collisions in Toronto (2017 - 2023) {#sec-types}
+When comparing the density of collisions in Toronto for each year with @fig-numcollisionmap, the density stays the same from 2017 to 2023 with ward 1 having the largest amount of collisions.
+According to @tbl-wardprofilestats and @sec-location, when comparing the the average household incomes for the wards each neighbourhood that make up 2% of motor vehicle collisions with the mean and median for average income per ward, the incomes are below the mean and median.
+The wards of the other 2 neighbourhoods from @tbl-wardprofilestats have incomes that are below the mean and either equal to or below the median.
+However for population sizes, the wards of West Humber-Clairville, York University Heights, Wexford/Maryvale, and Etobicoke City Centre are above the mean population size per ward while the rest in @tbl-wardprofilestats are below it.
+
+### Types of motor vehicle collisions in Toronto (2017 - 2023) {#sec-types}
 ```{r}
 #| label: tbl-collisiontype
-#| tbl-cap: Number of Motor Vehicle Collisions of Different Types Across Toronto from 2017 to 2023
+#| tbl-cap: Motor vehicle collisions of different types in Toronto from 2017 to 2023
 #| echo: false # Code will run and its output will show but the code itself will not be included in the doc
@@
   cleaned_collisions_data |>
+  # Group data by year
   group_by(year) |>
+  # Obtain unique rows 
   unique() |>
+  # Add in new column for proportions
   mutate(
@@
   ) |>
+  # Ungroup the data 
   ungroup() |>
-  rename (
+  # Rename columns 
+  rename(
     "Year" = year,
-    "Collision Type" = collision_type,
+    "Motor Vehicle Collision Type" = collision_type,
     "Number of Collisions" = num_of_collisions,
-    "Proportion of Collisions (Per Year)" = proportion
+    "Proportion of Motor Vehicle Collisions (Per Year)" = proportion
   ) |>
-  select("Year", "Collision Type", "Number of Collisions", "Proportion of Collisions (Per Year)")
+  # Choosing columns to display for table
+  select("Year", "Motor Vehicle Collision Type", "Number of Collisions", "Proportion of Motor Vehicle Collisions (Per Year)")
+
 cleaned_collisions_data_table |>
+  # Creating table 
   kable()
@@
 #| label: tbl-collisionstats
-#| tbl-cap: Summary Statistics of the Number of Motor Vehicle Collisions of Different Types Across Toronto from 2017 to 2023
+#| tbl-cap: Minimum, quartiles, median, and maximum of motor vehicle collisions of different types in Toronto from 2017 to 2023
 #| echo: false # Code will run and its output will show but the code itself will not be included in the doc
@@
 #| message: false # Messages are not included
+
+# Creates table
 cleaned_collisions_data_table |>
@@
 #| label: fig-collisionhistogram
-#| fig-cap: This histogram reveals the number of motor vehicle collisions from 2017 to 2023 across Toronto and the different types of motor vehicle collisions. 
+#| fig-cap:  Number of motor vehicle collisions in Toronto grouped by types from 2017 to 2023
 #| echo: false
@@
 #| message: false
-# Referenced chapter Chapter 5 of Telling Stories with Data by Rohan Alexander for 
-# scale_fill_viridis_d(), theme(), mapping = aes(...)
+
+# Referenced tutorial notes from summer STA238 2023 to obtain the code portion of geom_bar(stat="identity")
+# Code referenced: Alexander (2023) 
+# Creating histogram breaking down the types of motor vehicle collisions.
 cleaned_collisions_data |>
@@
 
-### Pedestrians Involved in a Motor Vehicle Collision (2017-2023) {#sec-pedes}
+### Pedestrians involved in a motor vehicle collision (2017-2023) {#sec-pedes}
 
@@
 #| label: tbl-pedestrianhistogram
-#| tbl-cap: Number of Pedestrians Involved in a Motor Vehicle Collision from 2017 to 2023
+#| tbl-cap: Number of pedestrians involved in motor vehicle collisions in Toronto from 2017 to 2023
 #| echo: false # Code will run and its output will show but the code itself will not be included in the doc
@@
 
-# Add up num of pedestrians for each year, get number of collisions each year, and the proportion
+# Setting up table 
 cleaned_collisions_data_yearly <-
   cleaned_collisions_data |>
+  # Groups data by year
   group_by(year) |>
+  # Obtains unique rows
   unique() |>
+  # Creates new columns for number of pedestrians involved in collisions for each year, gets the number of collisions for each year, and calculates the proportion of pedestrians involved in collisions each year
   mutate (
-    yearly_num_pedestrians = sum(num_of_pedestrians)
-  ) |>
-  mutate(
-    num_of_collisons_year = sum(num_of_collisions)
-  ) |>
-  mutate(
+    yearly_num_pedestrians = sum(num_of_pedestrians),
+    num_of_collisons_year = sum(num_of_collisions),
     proportion = round(yearly_num_pedestrians / num_of_collisons_year, 3)
@@
 
+# Creates table for pedestrians involved in collisions each year 
 cleaned_collisions_data_yearly |>
+  # Choose the column to display for the graph
   select(year, yearly_num_pedestrians, proportion) |>
+  # Filter out non-unique rows 
   unique() |>
+  # Rename variables to be clearer for table 
   rename(
@@
 #| label: fig-pedestrianhistogram
-#| fig-cap: This histogram reveals the number of pedestrians involved in a motor vehicle collisions from 2017 to 2023 across Toronto. 
+#| fig-cap: Number of pedestrians involved in motor vehicle collisions from 2017 to 2023 in Toronto. 
 #| echo: false
@@
 
-# Referenced tutorial notes from summer STA238 2023 to obtain the code portion of geom_bar(stat="identity") and plot.title
-# Referenced chapter Chapter 5 of Telling Stories with Data by Rohan Alexander for 
-# scale_fill_viridis_d(), theme(), mapping = aes(...)
-
+# Code referenced: Alexander (2023) 
+# Creates histogram for the number of pedestrians involved in motor vehicle collisions.
 cleaned_collisions_data_yearly |>
@@
 
-According to @tbl-wardprofilestats and @sec-location, ward 1 and ward 21's population number are above the mean of 110452 people (also known as average) however both wards are below the mean of $46,436 in terms of average household income.
+Vision Zero has also faced criticism due to its lack of focus on addressing road and pedestrian safety in low-income areas [@visionzeroinequality].
+While we did not directly looked at income compared to wards, which could be further investigated in the future, we found that the number of crashes across Toronto stayed pretty consistent from 2017 to 2023 based on @fig-numcollisionmap and the neighbourhoods we found impacted with high amounts of vehicle collisions in @tbl-neighbhourhoodtable were in wards that are below the mean average household income based on data from 2020.
 
@@
 #| label: tbl-wardprofile
-#| tbl-cap: Toronto Ward Profile in 2020
+#| tbl-cap: Population size and average household income across Toronto wards in 2020
 #| echo: false # Code will run and its output will show but the code itself will not be included in the doc
@@
     "Population Number" = pop_num, 
-    "Average Household Income"= avg_income
+    "Average Household Income (CAD)"= avg_income
   ) |>
@@
 
+```{r}
+#| label: tbl-neighbhourhoodtablefull
+#| tbl-cap: Toronto Neighbourhoods with the most number of motor vehicle collisions from 2017 to 2023
+#| echo: false # Code will run and its output will show but the code itself will not be included in the doc
+#| warning: false # Warnings are not included 
+#| message: false # Messages are not included 
+collisions_num_per_neighbhourhood |>
+  kable() 
+```
 \newpage

### 2024-01-26 outputs/paper/paper.qmd
Finish paper and updated wards profile dataset

@@
 # Results {#sec-result}
-### Motor vehicle collisions across Toronto neighbourhoods and wards (2017-2023) {#sec-location}
+## Motor vehicle collisions across Toronto neighbourhoods and wards (2017-2023) {#sec-location}
 As of 2024, There are 25 wards and 158 neighbourhoods in the city of Toronto.
@@
 
-### Types of motor vehicle collisions in Toronto (2017 - 2023) {#sec-types}
+## Types of motor vehicle collisions in Toronto (2017 - 2023) {#sec-types}
 ```{r}
@@
 ```
-From @fig-collisionhistogram, when comparing the overall number of collisions before 2020 and 2020 onwards, there is a overall annual decrease with the overall number of collisions.
+From @fig-collisionhistogram, when comparing the overall number of motor vehicle collisions before 2020 and 2020 onwards, there is a overall annual decrease with the number of collisions.
 However, collisions involving property damage remains the majority of collision cases over the years.
 We can see this decrease from 2019 and 2020 from about 79205 collisions to 43297 collisions.
-According to @tbl-collisionstats and @tbl-collisiontype, the peak number of property damage collisions was at 60865 in 2019, which made up 77% of the collisions in that year.
+According to @tbl-collisionstats and @tbl-collisiontype, the peak number of property damage collisions from 2017 to 2023 was at 60865 in 2019, which made up 77% of the collisions in that year.
 However property damage collisions in 2020 was up to 79% of collisions in 2020.
+We can also see that the lowest number of collisions occurred in 2021 and had stayed below 60865 since 2020.
+@fig-collisionhistogram and @tbl-collisiontype also reveals that 2020 to 2022, the second most common collision type was Personal injury (making up 10% to 12% of collisions in their respective years).
+In other years, the second most common motor vehicle collisions are Fail to Remain collisions.
 
-### Pedestrians involved in a motor vehicle collision (2017-2023) {#sec-pedes}
+## Pedestrians involved in a motor vehicle collision (2017-2023) {#sec-pedes}
 
@@
 ```
+From @fig-pedestrianhistogram, we can see that there has been a decline in pedestrian involvement in motor vehicle collisions from 2017 to 2023.
+Pedestrian involvement was at its highest point in 2017 with 1748 pedestrians where approximately 24% of the total number of collisions in 2017 involved pedestrians.
+In 2018, 1601 pedestrians were involved in motor vehicle collisions (approximately 21%).
+With 2019, 2020, and 2021, it was 1545 (approximately 20%), 1055 (approximately 24%), and 990 (approximately 24%), respectively.
+In 2022, the number of pedestrians increased to 1241 (approximately 22%) but fell to its lowest number of pedestrian in collisions in year with 968 people in 2023 (approximately 20%).
 
@@
 Vision Zero has also faced criticism due to its lack of focus on addressing road and pedestrian safety in low-income areas [@visionzeroinequality].
-While we did not directly looked at income compared to wards, which could be further investigated in the future, we found that the number of crashes across Toronto stayed pretty consistent from 2017 to 2023 based on @fig-numcollisionmap and the neighbourhoods we found impacted with high amounts of vehicle collisions in @tbl-neighbhourhoodtable were in wards that are below the mean average household income based on data from 2020.
+While we did not directly looked at income compared to wards, which could be further investigated in the future, we found that the number of crashes across Toronto stayed consistent from 2017 to 2023 based on @fig-numcollisionmap and the neighbourhoods we found impacted with high amounts of vehicle collisions in @tbl-neighbhourhoodtable were in wards that are below the mean average household income based on data from 2020.
 

### 2024-01-26 scripts/02-data_cleaning.R

@@ -179,7 +179,7 @@ ward_pop_size <-
 # Get row that contains the average household income
 ward_avg_income_row <- 
   cleaned_ward_data |> 
-  filter(x2021_one_variable_city_of_toronto_profiles == "Total - Household total income groups in 2020 for private households - 25% sample data")
+  filter(x2021_one_variable_city_of_toronto_profiles == "Average total income of households in 2020 ($)")
 # Convert row to a column 
 ward_avg_income <- 
   ward_avg_income_row |>

### 2024-01-26 outputs/paper/paper.qmd
Finalized paper, updated references, and updated README.md

@@
 date-format: "D MMMM YYYY"
-abstract: "As one of the fastest growing cities and densest cities in Canada, road and pedestrian safety are growing concerns in Toronto, especially after the COVID-19 Pandemic.
+abstract: "As one of the fastest-growing and densest Canadian cities, road and pedestrian safety are growing concerns in Toronto, especially after the COVID-19 pandemic.
 This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.
-The results shows that motor vehicle collisions and pedestrian involvement in them have decreased after the pandemic but are still prevalent in more lower-income areas.
-Further investigation is needed on the demographics of Toronto areas with high number of motor vehicle collisions."
+The results show that motor vehicle collisions and pedestrian involvement in them have decreased during and after the pandemic but are prevalent in the same areas from 2017 to 2023.
+Further investigation is needed on the demographics of Toronto areas with high numbers of motor vehicle collisions."
 format:
@@
 As of 2024, Toronto was reported to have one of the worst traffic congestion in North America [@citeWorstCongestion].
-As Toronto's population increases over the years, congestion on the road increased and road and pedestrian safety is a growing concern for many.
+As Toronto's population increases over the years, congestion on the road has increased and road and pedestrian safety is a growing concern for many.
 Some also believe that roads have gotten unsafer in the years after the pandemic.
-In a 2023 study done by Ipsos surveying Canadians on road safety, they mentioned that concerns about road safety has increased to 65% in 2023 from 59% in 2021 [@citeIpsos].
-In order to reduce traffic collision injuries and fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating fatalities on roads all together [@visionzero].
-An ambitious goal yet it raises the question how effective is this plan across Toronto over the years.
-We investigated this question and the following: how does motor vehicle collisions compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?
+In a 2023 study done by Ipsos surveying Canadians on road safety, they mentioned that concerns about road safety have increased to 65% in 2023 from 59% in 2021 [@citeIpsos].
+To reduce traffic collision injuries and fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating all fatalities on roads [@visionzero].
+An ambitious goal yet it raises the question of how effective is this plan across Toronto over the years.
+We investigated this question and the following: how do motor vehicle collisions compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?
 
@@
 In our findings, there was a decline in motor vehicle collisions overall from 2020 onwards.
-However, we saw that the locations motor vehicle collisions occurred at stayed consistent over the years.
-We also found that the wards that make up a greater proportion of motor vehicle collisions had average household incomes that are below Toronto's overall average household income.
-By understanding collision trends over the years in Toronto, it can aid with identifying areas of improvement of current safety plans like Vision Zero and overall helping others feel safer in the city.
+However, we saw that the locations where motor vehicle collisions occurred stayed consistent over the years.
+We also found that the majority of wards that make up a greater proportion of motor vehicle collisions had average household incomes that are below Toronto's overall average household income.
+Understanding collision trends over the years in Toronto can aid with identifying areas of improvement of current safety plans like Vision Zero and overall helping others feel safer in the city.
 
-In the rest of this paper, the data section will cover the datasets used, define variables used by our tables and graphs, and briefly explain the data cleaning process.
+In the rest of this paper, the data section will cover the datasets used, define variables used by our tables and graphs, and briefly explain the data-cleaning process.
 In the results section, we will reveal tables and graphs made on our datasets and explain what they show.
-In the discussion section, we will connect back to the real world and explain what the results could mean, the implication of our results, potential areas of improvement for the paper, and suggestions for future works.
+In the discussion section, we will connect back to the real world and explain what the results could mean, the implications of our results, potential areas of improvement for the paper, and suggestions for future works.
 Finally, the appendix section will include the entire table of some of our results.
@@
 # Data {#sec-data}
-The data used by the paper are `City Wards` [@citecitywards], `Ward Profiles (25-Ward Model)` [@citewardprofiles], and `Police Annual Statistical Report - Traffic Collisions` [@citecollisionsdata]  datasets, which were sourced from the City of Toronto's Open Data Portal.
+
+The data used by the paper are "City Wards" [@citecitywards], "Ward Profiles (25-Ward Model)" [@citewardprofiles], and "Police Annual Statistical Report - Traffic Collisions" [@citecollisionsdata]  datasets, which were sourced from the City of Toronto's Open Data Portal.
 These datasets were accessed using the `opendatatoronto` library [@citeopendatatoronto].
-Another dataset that could have been used was the `About Motor Vehicle Collisions involving Killed or Seriously Injured Persons` dataset from Open Data Toronto Portal.
+Another dataset that could have been used was the "About Motor Vehicle Collisions involving Killed or Seriously Injured Persons" dataset from Open Data Toronto Portal.
 However, this dataset's most current record was in 2022 and is outdated for our purposes.
@@
 The datasets were simulated, cleaned, analyzed, and tested using the `R` programming language [@citeR], `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
+The variables created for the datasets used was formed by grouping rows by existing columns and counting the number of rows with certain column variables.
+However, not all variables from our datasets were used or is visible on the graph.
+The original datasets was filtered for the years 2017 to 2023 and rows with missing information were omitted.
+These are the name of the following datasets with variables used for our analysis: cleaned_city_wards_data (`area_short_code`, `geometry`); cleaned_map_data (`year`, `collision_type`, `long`, `lat`, `neighbourhood`, `pedestrian_involved`, `num_of_collisions`, `yearly_collision_num`, `total_collisions_2017_2023`); cleaned_collisions_data (`year`, `collision_type`, `num_of_collisions`, `num_of_pedestrians`); cleaned_ward_data (`ward_num`, `ward_name`, `pop_num`,  `avg_income`).
+In our dataset, `year` represents a year from 2017 to 2023, inclusive.
+To view the cleaned datasets created, see @sec-datasetused under the Appendix section for more information.
 
-Some of the datasets used for the data analysis combined the `Ward Profiles (25-Ward Model)` and `Police Annual Statistical Report - Traffic Collisions` in order to create datasets with the following variables: `collision_type`, `num_of_collisions`, `num_of_pedestrians`, `yearly_collision_num`, and `total_collisions_2017_2023`.
-`collision`.
-These variables were formed by mostly created group existing columns and counting the number of rows with certain column variables.
+## Police Annual Statistical Report - Traffic Collisions
 The Toronto Police Services has defined motor vehicle collisions under 4 different categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata].
 Fatal means that a person was killed during the vehicle collision, personal injury indicates a person was injured during the collision, fail to remain means that the person left the scene before after information was provided regarding the collision, and property damage indicates that property has been damaged during the collision.
-With this, the 4 types are possible values under the variable `collision_type`.
-`num_of_collisions` represents the number of collisions that occurred in either a neighbhourhood in a specific year or the number of collisions of a certain type of a certain year depending on the context.
-Due to the complexity of motor vehicle collisions and the different combinations of ways they can be categorized, our analysis looked at collisions that were defined in only 1 of the 4 categories.
+The datasets, cleaned_collisions_data and cleaned_map_data, are cleaned versions of Police Annual Statistical Report - Traffic Collisions dataset and contains around 56 observations and 113,129 observations respectively.
+The
+cleaned_collisions_data datasets contains a summary of the number of collisions and pedestrians involved for each collision type for each year.
+On the other hand, cleaned_map_data contains more information regarding the neighbourhood where each collisions occurs.
 
-`num_of_pedestrians` represents the number of pedestrians that were hit in the vehicle collision.`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood mentioned in the `Police Annual Statistical Report - Traffic Collisions` dataset.
-For the dataset regarding Toronto ward profiles, the following variables were created: `ward_name` (indicates the name of the Toronto ward), `pop_num` (represents the population size of Toronto wards as of 2020), and `avg_income` (signifies the average household income of each ward as of 2020).
+The 4 types are possible values under the variable `collision_type`, which is also named "Type of Collision".
+Due to the various ways motor vehicle collisions can be categorized and the complexity of it, our analysis looked at motor vehicle collisions that were defined in only 1 of the 4 categories.
+`num_of_collisions` denotes the number of collisions that occurred in either a neighbhourhood in a specific year (also called "Number of Motor Vehicle Collisions") or the number of collisions of a certain type of a certain year depending on the context (named as "Number of Collisions").
+`num_of_pedestrians` represents the number of pedestrians that were hit in a vehicle collision in each collision type per year.
+`pedestrian_involved` signifies if a pedestrian was hit a collision with a "Yes" or "No".
+`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood.
+Three variables found in the cleaned_map_data `long`, `lat`, and `neighbourhood`, which represents longitude of collision site, latitude of collision site, and the neighbourhood name and number where collision site occurred in.
+
+## City Wards 
+The City Wards dataset, as known as cleaned_city_wards_data, contains that was used to recreate a map of Toronto with its ward divisions.
+The following variables were used in our analysis: `geometry` represents the coordinates of the ward divisions and `area_short_code` represents the ward number.
+
+## Ward Profiles (25-Ward Model)
+For the dataset regarding Toronto ward profiles known as cleaned_ward_data, the following variables were created: `ward_name` signifies the ward number, `ward_name` indicates the name of the Toronto ward, `pop_num` represents the population size of Toronto wards as of 2020, and `avg_income` denotes the average household income of each ward as of 2020.
+The dataset was extracted by selecting the rows "Average total income of households in 2020 ($)" and "Total - Age", where "Total - Age" contained the population size of each ward.
 
@@
 
-@tbl-wardprofilestats reveals that the mean (also known as the average) population size per ward in Toronto is 110452 people with a standard deviation of 10594 people and the mean for the average household income per ward is \$46,436 with a standard deviation of \$11,647.51.
-It also indicates that the median for the population size per ward is 110095 people and the median for average household income per ward is $45,345.
-From @tbl-wardprofile, ward 1's average household income is \$38,135 with a population of 115120 people and ward 21's average household income is \$40,565 with a population of 115120 people.
-For ward 14, 7, 20, and 2, the average households incomes are \$46,235, \$37,675, \$41,905, and \$45,345, respectively and the population sizes are 104555, 111200, 110095, and 117200 people, respectively.
+@tbl-wardprofilestats reveals that the mean (also known as the average) population size per ward in Toronto is 110452 people with a standard deviation of 10594 people and the mean for the average household income per ward is \$120,096 with a standard deviation of \$33,980.64.
+It also indicates that the median for the population size per ward is 110095 people and the median for average household income per ward is $107,300.
+From @tbl-wardprofile, ward 1's average household income is \$95,200 with a population of 115120 people and ward 21's average household income is \$91,500 with a population of 115120 people.
+For ward 14, 7, 20, and 2, the average households incomes are \$130,800, \$85,700, \$102,200, and \$146,600, respectively and the population sizes are 104555, 111200, 110095, and 117200 people, respectively.
 
 When comparing the density of collisions in Toronto for each year with @fig-numcollisionmap, the density stays the same from 2017 to 2023 with ward 1 having the largest amount of collisions.
-According to @tbl-wardprofilestats and @sec-location, when comparing the the average household incomes for the wards each neighbourhood that make up 2% of motor vehicle collisions with the mean and median for average income per ward, the incomes are below the mean and median.
-The wards of the other 2 neighbourhoods from @tbl-wardprofilestats have incomes that are below the mean and either equal to or below the median.
-However for population sizes, the wards of West Humber-Clairville, York University Heights, Wexford/Maryvale, and Etobicoke City Centre are above the mean population size per ward while the rest in @tbl-wardprofilestats are below it.
+According to @tbl-neighbhourhoodtable, @tbl-wardprofilestats, @fig-numcollisionmap, and @citefinder, the average household incomes for 3 out of 4 of the wards of each neighbourhood that make up 2% of motor vehicle collisions are below the mean and median for average income per ward with their population size being higher than the mean population size per ward.
+1 ward in the 2% has an average household income above the mean and median for average income per ward and their population size is below the mean and median population size per ward.
+The wards of the other 2 neighbourhoods from @tbl-neighbhourhoodtable, Clairlea-Birchmount's average household income is below the mean and median with its population size being above or equal to the mean and median.
+On the other hand, Etobicoke City Centre's average household income is above the mean and median and the population size is above the mean and median.
 
@@
   ggtitle("Number of Motor Vehicle Collisions in Toronto from 2017 to 2023") +
-  labs(x = "Year", y = "Number of Collisions", fill = "Type of Collisions") +
+  labs(x = "Year", y = "Number of Collision", fill = "Type of Collisions") +
   theme(legend.position = "bottom", plot.title = element_text(size = 9, face = "bold"), rect = element_blank(), legend.title = element_text(size = 9), legend.text = element_text(size = 7)) +
@@
 From @fig-collisionhistogram, when comparing the overall number of motor vehicle collisions before 2020 and 2020 onwards, there is a overall annual decrease with the number of collisions.
-However, collisions involving property damage remains the majority of collision cases over the years.
+The result also reveals that the number of motor vehicle collisions increased from 2017 to 2019 before decreasing in 2020.
+Collisions involving property damage remains the majority of collision cases over the years.
 We can see this decrease from 2019 and 2020 from about 79205 collisions to 43297 collisions.
@@
 In other years, the second most common motor vehicle collisions are Fail to Remain collisions.
+
+\newpage
 
@@
 ```
-
+\newpage
 ```{r}
@@
 # Discussion
+From @sec-result, we found that from 2017 to 2023, the number of motor vehicle collisions has decreased during 2020 and onwards, the number of pedestrians involved in these collisions has decreased, and that hotspots for motor vehicle collisions has stayed consistent.
+We also found that the most common type of collisions involved property damage and the second most common collision type in 2020 to 2022 was Personal injury and for other years it was Fail to Remain.
+We also observed that there was a decrease in numbers from 2019 to 2020 such as the number of pedestrians involved in collisions in Toronto decreased by approximately 38% and the number of collisions in Toronto decreased by 59%.
+Our observations also revealed that the number of pedestrians and number of collisions were below the numbers reached in the years prior to 2020, which was when the COVID-19 pandemic started.
 
-## First discussion point {#sec-first-point}
+Throughout 2020 to 2021, the Ontario Government imposed Stay-at-Home orders, requiring Canadians to stay at home unless they are travelling for essential purposes such as grocery shopping [@citeStayAtHome].
+This could have been one of the reasons why there was a decrease in motor vehicle collisions by 59% and 2020 and 2021 having fewer number of accidents.
+In 2022, the Ontario Government began plans to reopen Ontario in the beginning of the year, which led to more people on the road and outside and therefore could have resulted in the number of collisions in 2022 being higher than in 2020 and 2021 [@citeReopening].
 
+Vision Zero Safety Plan was a plan created by the City of Toronto in 2016 to reduce traffic collision injuries and fatalities [@visionzero].
+Based on the decreasing number of pedestrians being involved in collisions over the years, it suggests Vision Zero could have had an influence on this trend.
+However, the number of motor vehicle collisions increased from 2017 to 2019, which means there are other factors in play that could be further investigated by future works.
 Vision Zero has also faced criticism due to its lack of focus on addressing road and pedestrian safety in low-income areas [@visionzeroinequality].
-While we did not directly looked at income compared to wards, which could be further investigated in the future, we found that the number of crashes across Toronto stayed consistent from 2017 to 2023 based on @fig-numcollisionmap and the neighbourhoods we found impacted with high amounts of vehicle collisions in @tbl-neighbhourhoodtable were in wards that are below the mean average household income based on data from 2020.
+While we did not directly looked at the relationship of ward income and number of collisions, which could be investigated in future works, we found that the number of crashes across Toronto stayed consistent from 2017 to 2023 based on @fig-numcollisionmap.
+The neighbourhoods we found impacted with high amounts of vehicle collisions (composing approximately 2% of total collisions) in @tbl-neighbhourhoodtable were mostly in wards that are below the mean average household income based on data from 2020.
 
-## Second discussion point
+Despite using datasets with over 100,000 observations, the datasets are not 100% accurate.
+In data entry, human error has occurred.
+This is seen with the original dataset containing rows that are partially completed with N/A as some of their entries.
+As mentioned by @citecollisionsdata, in order to ensure privacy, coordinates to the collision sites were offset to the closest road intersection.
+This means that the neighbourhoods where collisions were recorded at might not be completely accurate.
+Other errors that may have occurred includes motor vehicles collisions being purposely misrepresented when entered or inaccurate information was entered by mistake.
 
-## Third discussion point
+The analysis was faced with several limitations as well.
+The data on Toronto Wards' average incomes and population size was done in 2020, which may not reflect trends of other years.
+Due to the complexity of motor vehicle collisions and the different they can be categorized, our analysis looked at collisions that were defined in only 1 of the 4 categories (Fatal, Fail to Remain, Property Damage, and Personal Injury).
+This means that numbers of motor vehicle collisions and number of pedestrians involvement with them are lower than they would be.
+Since no dataset currently exists that corresponds neighbourhoods with their respective wards, searches for each neighbourhood and their corresponding ward was done manually using the Toronto's Find Your Neighbourhood page [@citefinder].
 
-## Weaknesses and next steps
+Future investigations could be conduct on the demographics of Toronto locations with high number of motor vehicle collisions.
+Other investigations that could be made includes ones regarding the increase in motor vehicle collisions from 2017 to 2019 and the relationship of ward income and number of collisions.
+Datasets could also be created that maps Toronto neighbourhoods to their respective wards to allow for more in-depth and accurate analysis of motor vehicle collisions in different neighbourhoods and wards.
 
@@
 # Appendix {-}
+
+## Datasets Used {#sec-datasetused}
+Due to the length of the datasets horizontally, presenting glimpses of the datasets was not possible in the paper.
+However, the cleaned datasets that was used can be found here: https://github.com/moonsdust/toronto-collisions/tree/main/outputs/data
+
+## Tables from Results {#sec-tablesfromresults}
 ```{r}

### 2024-01-26 scripts/02-data_cleaning.R

@@ -139,6 +139,8 @@ ward_pop_size_row <-
 # Convert row to a column 
 ward_pop_size <- 
   ward_pop_size_row |>
+  # Code snippet to convert to a 1-column matrix and then to a vector 
+  # was adapted from the following: https://www.geeksforgeeks.org/convert-matrix-to-vector-in-r/
   t() |> # Convert 1-row matrix into a 1-column matrix 
   c() |> # Convert to a vector
   tibble() |> # Change into a tibble

### 2024-01-26 outputs/paper/paper.qmd
updated paper

@@
 abstract: "As one of the fastest-growing and densest Canadian cities, road and pedestrian safety are growing concerns in Toronto, especially after the COVID-19 pandemic.
-This paper looks at trends of collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.
+This paper looks at trends of motor vehicle collisions from 2017 to 2023 in Toronto neighbourhoods and wards, types of collisions, and the number of pedestrians involved.
 The results show that motor vehicle collisions and pedestrian involvement in them have decreased during and after the pandemic but are prevalent in the same areas from 2017 to 2023.
-Further investigation is needed on the demographics of Toronto areas with high numbers of motor vehicle collisions."
+Further investigation is needed on the demographics of Toronto areas with a high number of motor vehicle collisions."
 format:
@@
 To reduce traffic collision injuries and fatalities, the city of Toronto created the Vision Zero Safety Plan in 2016, which contains over 50 safety measures with the ultimate goal of eliminating all fatalities on roads [@visionzero].
-An ambitious goal yet it raises the question of how effective is this plan across Toronto over the years.
+It's an ambitious goal yet it raises the question of how effective this plan across Toronto over the years.
 We investigated this question and the following: how do motor vehicle collisions compare from the early years of the plan to the present following the outbreak of the COVID-19 pandemic in 2020?
 
-In this paper, collisions trends were analyzed from 2017 to 2023 specifically into motor vehicle collisions in Toronto neighbourhoods and wards, the types of motor vehicle collisions that occurred, and the number of pedestrians involved in them.
+In this paper, collision trends were analyzed from 2017 to 2023, specifically into motor vehicle collisions in Toronto neighbourhoods and wards, the types of motor vehicle collisions that had occurred, and the number of pedestrians involved in them.
 In our findings, there was a decline in motor vehicle collisions overall from 2020 onwards.
-However, we saw that the locations where motor vehicle collisions occurred stayed consistent over the years.
-We also found that the majority of wards that make up a greater proportion of motor vehicle collisions had average household incomes that are below Toronto's overall average household income.
-Understanding collision trends over the years in Toronto can aid with identifying areas of improvement of current safety plans like Vision Zero and overall helping others feel safer in the city.
+However, we saw that the locations where motor vehicle collisions had occurred stayed consistent over the years.
+We also found that the majority of wards that make up a larger proportion of motor vehicle collisions had average household incomes that are below Toronto's overall average household income.
+Understanding collision trends over the years in Toronto can aid with identifying areas of improvement of current safety plans like Vision Zero and overall help others feel safer in the city.
 
@@
 In the discussion section, we will connect back to the real world and explain what the results could mean, the implications of our results, potential areas of improvement for the paper, and suggestions for future works.
-Finally, the appendix section will include the entire table of some of our results.
+Finally, the appendix section will include the entire table of some of our results and other information.
 
@@
 These datasets were accessed using the `opendatatoronto` library [@citeopendatatoronto].
-Another dataset that could have been used was the "About Motor Vehicle Collisions involving Killed or Seriously Injured Persons" dataset from Open Data Toronto Portal.
+Another dataset that could have been used was the "About Motor Vehicle Collisions involving Killed or Seriously Injured Persons" dataset from the City of Toronto's Open Data Portal.
 However, this dataset's most current record was in 2022 and is outdated for our purposes.
@@
 The datasets were simulated, cleaned, analyzed, and tested using the `R` programming language [@citeR], `tidyverse` [@citetidyverse], `knitr` [@citeknitr], `janitor`[@citejanitor], `dplyr` [@citedplyr], `ggplot2` [@citeggplot2], and `sf` [@citesf].
-The variables created for the datasets used was formed by grouping rows by existing columns and counting the number of rows with certain column variables.
-However, not all variables from our datasets were used or is visible on the graph.
-The original datasets was filtered for the years 2017 to 2023 and rows with missing information were omitted.
-These are the name of the following datasets with variables used for our analysis: cleaned_city_wards_data (`area_short_code`, `geometry`); cleaned_map_data (`year`, `collision_type`, `long`, `lat`, `neighbourhood`, `pedestrian_involved`, `num_of_collisions`, `yearly_collision_num`, `total_collisions_2017_2023`); cleaned_collisions_data (`year`, `collision_type`, `num_of_collisions`, `num_of_pedestrians`); cleaned_ward_data (`ward_num`, `ward_name`, `pop_num`,  `avg_income`).
+The variables created for the datasets were formed by grouping rows by existing columns and counting the number of rows with certain column variables.
+However, not all variables from our datasets were used or are visible on the graph.
+The original datasets were filtered for the years 2017 to 2023 and rows with missing information were omitted.
+These are the name of the following datasets with variables used for our analysis: "cleaned_city_wards_data" (`area_short_code`, `geometry`); "cleaned_map_data" (`year`, `collision_type`, `long`, `lat`, `neighbourhood`, `pedestrian_involved`, `num_of_collisions`, `yearly_collision_num`, `total_collisions_2017_2023`); "cleaned_collisions_data" (`year`, `collision_type`, `num_of_collisions`, `num_of_pedestrians`); "cleaned_ward_data" (`ward_num`, `ward_name`, `pop_num`,  `avg_income`).
 In our dataset, `year` represents a year from 2017 to 2023, inclusive.
@@
 ## Police Annual Statistical Report - Traffic Collisions
-The Toronto Police Services has defined motor vehicle collisions under 4 different categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata].
-Fatal means that a person was killed during the vehicle collision, personal injury indicates a person was injured during the collision, fail to remain means that the person left the scene before after information was provided regarding the collision, and property damage indicates that property has been damaged during the collision.
-The datasets, cleaned_collisions_data and cleaned_map_data, are cleaned versions of Police Annual Statistical Report - Traffic Collisions dataset and contains around 56 observations and 113,129 observations respectively.
-The
-cleaned_collisions_data datasets contains a summary of the number of collisions and pedestrians involved for each collision type for each year.
-On the other hand, cleaned_map_data contains more information regarding the neighbourhood where each collisions occurs.
 
-The 4 types are possible values under the variable `collision_type`, which is also named "Type of Collision".
-Due to the various ways motor vehicle collisions can be categorized and the complexity of it, our analysis looked at motor vehicle collisions that were defined in only 1 of the 4 categories.
-`num_of_collisions` denotes the number of collisions that occurred in either a neighbhourhood in a specific year (also called "Number of Motor Vehicle Collisions") or the number of collisions of a certain type of a certain year depending on the context (named as "Number of Collisions").
+The datasets, "cleaned_collisions_data" and "cleaned_map_data" are clean versions of the Police Annual Statistical Report - Traffic Collisions dataset and contain around 56 observations and 113,129 observations, respectively.
+The "cleaned_collisions_data" dataset contains a summary of the number of collisions and pedestrians involved for each collision type for each year.
+On the other hand, "cleaned_map_data" contains more information regarding the neighbourhoods where each collision occurs.
+The Toronto Police Services has defined motor vehicle collisions under four categories: Fatal, Personal Injury, Fail to Remain, and Property Damage [@citecollisionsdata].
+Fatal means that a person was killed during the vehicle collision, Personal Injury indicates a person was injured during the collision, Fail to Remain means that the person left the scene before information was provided, and Property Damage indicates that property has been damaged during the collision.
+With this, they are the four possible values under the variable `collision_type`, which is also named "Type of Collision".
+Due to the various ways motor vehicle collisions can be categorized and their complexity, our analysis looked at motor vehicle collisions that were defined in only one of the four categories.
+
+`num_of_collisions` denotes the number of collisions that occurred in either a neighbourhood in a specific year (also called "Number of Motor Vehicle Collisions") or the number of collisions of a certain type of a certain year (named "Number of Collisions").
 `num_of_pedestrians` represents the number of pedestrians that were hit in a vehicle collision in each collision type per year.
 `pedestrian_involved` signifies if a pedestrian was hit a collision with a "Yes" or "No".
-`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbhourhood.
-Three variables found in the cleaned_map_data `long`, `lat`, and `neighbourhood`, which represents longitude of collision site, latitude of collision site, and the neighbourhood name and number where collision site occurred in.
+`yearly_collision_num` and `total_collisions_2017_2023` are the count of collisions in a specific year and the total number of collisions from 2017 to 2023 for each Toronto neighbourhood, respectively.
+Three variables are found in the cleaned_map_data `long`, `lat`, and `neighbourhood`, which represent the longitude of the collision site, the latitude of the collision site, and the neighbourhood name and number where the collision site occurred, respectively.
 
 ## City Wards 
-The City Wards dataset, as known as cleaned_city_wards_data, contains that was used to recreate a map of Toronto with its ward divisions.
+The City Wards dataset, also known as "cleaned_city_wards_data", contains information used to recreate a map of Toronto with its ward divisions.
+This dataset contains around 25 observations.
 The following variables were used in our analysis: `geometry` represents the coordinates of the ward divisions and `area_short_code` represents the ward number.
@@
 ## Ward Profiles (25-Ward Model)
-For the dataset regarding Toronto ward profiles known as cleaned_ward_data, the following variables were created: `ward_name` signifies the ward number, `ward_name` indicates the name of the Toronto ward, `pop_num` represents the population size of Toronto wards as of 2020, and `avg_income` denotes the average household income of each ward as of 2020.
-The dataset was extracted by selecting the rows "Average total income of households in 2020 ($)" and "Total - Age", where "Total - Age" contained the population size of each ward.
+The Ward Profiles dataset, also known as "cleaned_ward_data", has 25 observations and contains census data of Toronto's population such as income, population size, etc.
+For the variables in the dataset, `ward_name` signifies the ward number, `ward_name` indicates the name of the Toronto ward, `pop_num` represents the population size of Toronto wards as of 2020, and `avg_income` denotes the average household income of each ward as of 2020.
+The cleaned dataset selected columns from the original dataset by selecting the rows "Average total income of households in 2020 ($)" and "Total - Age", where "Total - Age" contains the population size of each ward.
 
@@
 ## Motor vehicle collisions across Toronto neighbourhoods and wards (2017-2023) {#sec-location}
-As of 2024, There are 25 wards and 158 neighbourhoods in the city of Toronto.
+As of 2024, there are 25 wards and 158 neighbourhoods in Toronto.
 The distribution of motor vehicle collisions across different neighbourhoods and wards was investigated and the following was obtained:
@@
   ) |>
+  # Rename columns
   rename(
@@
   select(pop_num, avg_income) |>
+  # Rename columns
   rename (
@@
 ```
-@tbl-neighbhourhoodtable reveals that the neighbourhood with the highest number of motor vehicle collisions and the approximately how many motor vehicle collisions from 2017 to 2023 occurred in each neighbourhood.
+@tbl-neighbhourhoodtable reveals the neighbourhood with the highest number of motor vehicle collisions and approximately how many motor vehicle collisions from 2017 to 2023 occurred in each neighbourhood.
 The neighbourhoods with a high number of collisions are the following: West Humber-Clairville (approximately 2%), Wexford/Maryvale (approximately 2%), South Riverdale (approximately 2%), York University Heights (approximately 2%), Clairlea-Birchmount (approximately 2%), Birchmount (approximately 1%), and Etobicoke City Centre (approximately 1%).
-The full table of @tbl-neighbhourhoodtable can be found here @tbl-neighbhourhoodtablefull in the Appendix section with the rest of the neighbourhoods making up approximately 1% of collisions or less.
+The full table of @tbl-neighbhourhoodtable can be found here @tbl-neighbhourhoodtablefull in the Appendix section with the rest of the neighbourhoods making up roughly 1% of collisions or less.
 
 In @tbl-neighbhourhoodtable, West Humber-Clairville has the largest amount of collisions with 2812 motor vehicle collisions from 2017 to 2023.
-West Humber-Clairville is located in ward 1, which from @fig-numcollisionmap is the region with the darkests shade of blue and has the highest density of points [@citefinder].
+West Humber-Clairville is located in ward 1, which from @fig-numcollisionmap is the region with the darkest shade of blue and has the highest density of points [@citefinder].
 Following this is Wexford/Maryvale, which is located in ward 21 with 2075 collisions [@citefinder].
-The rest of the neighbourhood are in the following wards based on @citefinder: South Riverdale is in Ward 14, York University Heights is in Ward 7, Clairlea-Birchmount is in Ward 20, and Etobicoke City Centre is in Ward 2.
+The rest of the neighbourhood is in the following wards based on @citefinder: South Riverdale is in Ward 14, York University Heights is in Ward 7, Clairlea-Birchmount is in Ward 20, and Etobicoke City Centre is in Ward 2.
 
+We will first define the following: mean means the average value amongst the dataset, median means the data point found in the middle of a dataset, and the standard deviation is how far are values from the mean of a dataset.
 @tbl-wardprofilestats reveals that the mean (also known as the average) population size per ward in Toronto is 110452 people with a standard deviation of 10594 people and the mean for the average household income per ward is \$120,096 with a standard deviation of \$33,980.64.
-It also indicates that the median for the population size per ward is 110095 people and the median for average household income per ward is $107,300.
-From @tbl-wardprofile, ward 1's average household income is \$95,200 with a population of 115120 people and ward 21's average household income is \$91,500 with a population of 115120 people.
-For ward 14, 7, 20, and 2, the average households incomes are \$130,800, \$85,700, \$102,200, and \$146,600, respectively and the population sizes are 104555, 111200, 110095, and 117200 people, respectively.
+It also indicates that the median population size per ward is 110095 people and the median average household income per ward is $107,300.
+From @tbl-wardprofile, Ward 1's average household income is \$95,200 with a population of 115120 people and Ward 21's average household income is \$91,500 with a population of 115120 people.
+For wards 14, 7, 20, and 2, the average households incomes are \$130,800, \$85,700, \$102,200, and \$146,600, respectively and the population sizes are 104555, 111200, 110095, and 117200 people, respectively.
 
 When comparing the density of collisions in Toronto for each year with @fig-numcollisionmap, the density stays the same from 2017 to 2023 with ward 1 having the largest amount of collisions.
-According to @tbl-neighbhourhoodtable, @tbl-wardprofilestats, @fig-numcollisionmap, and @citefinder, the average household incomes for 3 out of 4 of the wards of each neighbourhood that make up 2% of motor vehicle collisions are below the mean and median for average income per ward with their population size being higher than the mean population size per ward.
-1 ward in the 2% has an average household income above the mean and median for average income per ward and their population size is below the mean and median population size per ward.
-The wards of the other 2 neighbourhoods from @tbl-neighbhourhoodtable, Clairlea-Birchmount's average household income is below the mean and median with its population size being above or equal to the mean and median.
+According to @tbl-neighbhourhoodtable, @tbl-wardprofilestats, @fig-numcollisionmap, and @citefinder, the average household incomes for three out of four of the wards of each neighbourhood that make up 2% of motor vehicle collisions are below the mean and median average income per ward with their population size being higher than the mean population size per ward.
+One ward in the 2% has an average household income above the mean and median for average income per ward and its population size is below the mean and median population size per ward.
+With the wards of the other two neighbourhoods from @tbl-neighbhourhoodtable, Clairlea-Birchmount's average household income is below the mean and median with its population size being above or equal to the mean and median.
 On the other hand, Etobicoke City Centre's average household income is above the mean and median and the population size is above the mean and median.
@@
   ggtitle("Number of Motor Vehicle Collisions in Toronto from 2017 to 2023") +
-  labs(x = "Year", y = "Number of Collision", fill = "Type of Collisions") +
+  labs(x = "Year", y = "Number of Collision", fill = "Type of Collision") +
   theme(legend.position = "bottom", plot.title = element_text(size = 9, face = "bold"), rect = element_blank(), legend.title = element_text(size = 9), legend.text = element_text(size = 7)) +
@@
 ```
-From @fig-collisionhistogram, when comparing the overall number of motor vehicle collisions before 2020 and 2020 onwards, there is a overall annual decrease with the number of collisions.
+From @fig-collisionhistogram and @tbl-collisiontype, when comparing the overall number of motor vehicle collisions before 2020 and 2020 onwards, there is an overall annual decrease in the number of collisions.
 The result also reveals that the number of motor vehicle collisions increased from 2017 to 2019 before decreasing in 2020.
-Collisions involving property damage remains the majority of collision cases over the years.
+Collisions involving property damage remain the majority of collision cases over the years.
 We can see this decrease from 2019 and 2020 from about 79205 collisions to 43297 collisions.
-According to @tbl-collisionstats and @tbl-collisiontype, the peak number of property damage collisions from 2017 to 2023 was at 60865 in 2019, which made up 77% of the collisions in that year.
-However property damage collisions in 2020 was up to 79% of collisions in 2020.
-We can also see that the lowest number of collisions occurred in 2021 and had stayed below 60865 since 2020.
-@fig-collisionhistogram and @tbl-collisiontype also reveals that 2020 to 2022, the second most common collision type was Personal injury (making up 10% to 12% of collisions in their respective years).
+According to @tbl-collisionstats and @tbl-collisiontype, the highest number of property damage collisions from 2017 to 2023 was at 60865 in 2019, which made up around 77% of the collisions in that year.
+However, property damage collisions in 2020 were up to 79% of collisions in 2020.
+We can also see that the lowest number of collisions occurred in 2021 and has stayed below 60865 since 2020.
+@fig-collisionhistogram and @tbl-collisiontype also reveal that from 2020 to 2022, the second most common collision type was Personal injury (making up 10% to 12% of collisions in their respective years).
 In other years, the second most common motor vehicle collisions are Fail to Remain collisions.
@@
 \newpage
+
 ```{r}
@@
 From @fig-pedestrianhistogram, we can see that there has been a decline in pedestrian involvement in motor vehicle collisions from 2017 to 2023.
-Pedestrian involvement was at its highest point in 2017 with 1748 pedestrians where approximately 24% of the total number of collisions in 2017 involved pedestrians.
+Pedestrian involvement was at its highest point in 2017 with 1748 pedestrians, where approximately 24% of the total number of collisions in 2017 involved pedestrians.
 In 2018, 1601 pedestrians were involved in motor vehicle collisions (approximately 21%).
-With 2019, 2020, and 2021, it was 1545 (approximately 20%), 1055 (approximately 24%), and 990 (approximately 24%), respectively.
-In 2022, the number of pedestrians increased to 1241 (approximately 22%) but fell to its lowest number of pedestrian in collisions in year with 968 people in 2023 (approximately 20%).
+In 2019, 2020, and 2021, it was 1545 (approximately 20%), 1055 (approximately 24%), and 990 (approximately 24%), respectively.
+In 2022, the number of pedestrians increased to 1241 (approximately 22%) but fell to its lowest number of pedestrians in collisions in a year with 968 people in 2023 (approximately 20%).
 
@@
 # Discussion
-From @sec-result, we found that from 2017 to 2023, the number of motor vehicle collisions has decreased during 2020 and onwards, the number of pedestrians involved in these collisions has decreased, and that hotspots for motor vehicle collisions has stayed consistent.
-We also found that the most common type of collisions involved property damage and the second most common collision type in 2020 to 2022 was Personal injury and for other years it was Fail to Remain.
-We also observed that there was a decrease in numbers from 2019 to 2020 such as the number of pedestrians involved in collisions in Toronto decreased by approximately 38% and the number of collisions in Toronto decreased by 59%.
-Our observations also revealed that the number of pedestrians and number of collisions were below the numbers reached in the years prior to 2020, which was when the COVID-19 pandemic started.
+From @sec-result, we found that from 2017 to 2023, the number of motor vehicle collisions decreased during 2020 and onwards, the number of pedestrians involved in these collisions decreased, and hotspots for motor vehicle collisions stayed consistent.
+We also found that the most common type of collision was Property Damage and the second most common collision type in 2020 to 2022 was Personal Injury and for other years it was Fail to Remain.
+We also observed that there was a decrease in numbers from 2019 to 2020 where the number of pedestrians involved in collisions in Toronto decreased by approximately 38% and the number of collisions in Toronto decreased by 59%.
+Our observations also revealed that the number of pedestrians and the number of collisions were below the numbers reached in the years prior to 2020, which was when the COVID-19 pandemic started.
 
-Throughout 2020 to 2021, the Ontario Government imposed Stay-at-Home orders, requiring Canadians to stay at home unless they are travelling for essential purposes such as grocery shopping [@citeStayAtHome].
-This could have been one of the reasons why there was a decrease in motor vehicle collisions by 59% and 2020 and 2021 having fewer number of accidents.
-In 2022, the Ontario Government began plans to reopen Ontario in the beginning of the year, which led to more people on the road and outside and therefore could have resulted in the number of collisions in 2022 being higher than in 2020 and 2021 [@citeReopening].
+Throughout 2020 to 2021, the Ontario Government imposed Stay-at-Home orders and required Canadians to stay at home unless they are travelling for essential purposes such as grocery shopping [@citeStayAtHome].
+This could have been one of the reasons why there was a decrease in motor vehicle collisions by 59% and 2020 and 2021 had fewer accidents.
+In 2022, the Ontario Government began plans to reopen Ontario at the beginning of the year, which led to more people on the road and being outside and therefore could have resulted in the number of collisions in 2022 being higher than in 2020 and 2021 in @fig-collisionhistogram [@citeReopening].
 
-Vision Zero Safety Plan was a plan created by the City of Toronto in 2016 to reduce traffic collision injuries and fatalities [@visionzero].
-Based on the decreasing number of pedestrians being involved in collisions over the years, it suggests Vision Zero could have had an influence on this trend.
-However, the number of motor vehicle collisions increased from 2017 to 2019, which means there are other factors in play that could be further investigated by future works.
+Vision Zero Safety Plan is a plan created by the City of Toronto in 2016 to reduce traffic collision injuries and fatalities [@visionzero].
+Based on the decreasing number of pedestrians being involved in collisions over the years, it suggests Vision Zero could have influenced this trend.
+However, the number of motor vehicle collisions increased from 2017 to 2019, which means other factors are in play that could be further investigated by future works.
 Vision Zero has also faced criticism due to its lack of focus on addressing road and pedestrian safety in low-income areas [@visionzeroinequality].
-While we did not directly looked at the relationship of ward income and number of collisions, which could be investigated in future works, we found that the number of crashes across Toronto stayed consistent from 2017 to 2023 based on @fig-numcollisionmap.
-The neighbourhoods we found impacted with high amounts of vehicle collisions (composing approximately 2% of total collisions) in @tbl-neighbhourhoodtable were mostly in wards that are below the mean average household income based on data from 2020.
+While we did not directly look at the relationship between ward income and the number of collisions, which could be investigated in future works, we found that the number of crashes across Toronto stayed consistent from 2017 to 2023 based on @fig-numcollisionmap.
+The neighbourhoods we found impacted with high amounts of vehicle collisions (composing about 2% of total collisions) in @tbl-neighbhourhoodtable were mostly in wards that are below the mean average household income based on data from 2020.
 
-Despite using datasets with over 100,000 observations, the datasets are not 100% accurate.
-In data entry, human error has occurred.
+Despite using datasets with over 100,000 observations, the datasets are not completely accurate.
+With the data entry for one of datasets, human error has occurred.
 This is seen with the original dataset containing rows that are partially completed with N/A as some of their entries.
@@
 This means that the neighbourhoods where collisions were recorded at might not be completely accurate.
-Other errors that may have occurred includes motor vehicles collisions being purposely misrepresented when entered or inaccurate information was entered by mistake.
+Other errors that may have occurred include motor vehicle collisions being purposely misrepresented when entered or inaccurate information being entered by mistake.
 
@@
 The data on Toronto Wards' average incomes and population size was done in 2020, which may not reflect trends of other years.
-Due to the complexity of motor vehicle collisions and the different they can be categorized, our analysis looked at collisions that were defined in only 1 of the 4 categories (Fatal, Fail to Remain, Property Damage, and Personal Injury).
-This means that numbers of motor vehicle collisions and number of pedestrians involvement with them are lower than they would be.
-Since no dataset currently exists that corresponds neighbourhoods with their respective wards, searches for each neighbourhood and their corresponding ward was done manually using the Toronto's Find Your Neighbourhood page [@citefinder].
+Due to the complexity of motor vehicle collisions and the different ways they can be categorized, our analysis looked at collisions that were defined in only one of the four categories (Fatal, Fail to Remain, Property Damage, and Personal Injury).
+This means that the number of motor vehicle collisions and the number of pedestrians involved with them are lower than they would be.
+Currently, no dataset exists that corresponds neighbourhoods with their respective wards and so searches for each neighbourhood and their ward were done manually using Toronto's Find Your Neighbourhood page [@citefinder].
 
-Future investigations could be conduct on the demographics of Toronto locations with high number of motor vehicle collisions.
-Other investigations that could be made includes ones regarding the increase in motor vehicle collisions from 2017 to 2019 and the relationship of ward income and number of collisions.
-Datasets could also be created that maps Toronto neighbourhoods to their respective wards to allow for more in-depth and accurate analysis of motor vehicle collisions in different neighbourhoods and wards.
-
+Future investigations could be conducted on the demographics of Toronto locations with a high number of motor vehicle collisions.
+Other investigations that could be made include ones regarding the increase in motor vehicle collisions from 2017 to 2019 and the relationship between ward income and the number of collisions.
+Datasets could also be created that map Toronto neighbourhoods to their respective wards to allow for more in-depth and accurate analysis of motor vehicle collisions in different neighbourhoods and wards.
 \newpage
@@
 
+## Acknowledgments 
+We would like the acknowledge @citetellingstorieswithdata for some of the R code used in this paper to produce the tables and graphs.
+
 ## Datasets Used {#sec-datasetused}
 Due to the length of the datasets horizontally, presenting glimpses of the datasets was not possible in the paper.
-However, the cleaned datasets that was used can be found here: https://github.com/moonsdust/toronto-collisions/tree/main/outputs/data
+However, the cleaned datasets that was used can be found here: https://github.com/moonsdust/toronto-collisions/tree/main/outputs/data.
 
